---
title: "TUTORIAL: One dataset, four meta-analyses: synthesising mean effects, within-population variability, and between-population heterogeneity in ecology"
date: "`r Sys.Date()`"
# author: "AM, CW, HL, ML, SN"
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 5
    toc-float: false
    toc-title: "**CONTENTS**"
    output-file: "index.html"
    embed-resources: true
    code-tools: 
      source: false
    number-sections: false
    page-layout: full
    code-overflow: wrap
    df_print: paged
    theme: yeti
    code-line-numbers: false
    # grid:
    #   body-width: 1400px 
    # 　margin-width: 180px
crossref: 
  fig-title: Figure
  tbl-title: Table
  title-delim: "—"
  fig-prefix: Fig.
  tbl-prefix: Table.
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
editor: 
  markdown: 
    wrap: sentence
---

<!-- Citation info!! -->

# Intended audience and scope of this tutorial

This tutorial is intended for researchers who already have a basic understanding of meta-analysis and some familiarity with statistical modelling in `R`.
It is aimed at readers who wish to develop practical skills in location–scale meta-analytic modelling, using standard meta-analytic models as a foundation for understanding how mean effects and variability can be jointly analysed.

The scope of this tutorial is deliberately focused on conceptual comparison and practical illustration rather than exhaustive coverage.
We do not provide an introduction to meta-analytic theory, effect size calculation, or general R programming, as these topics are well covered elsewhere.
Instead, the tutorial uses a single dataset to demonstrate four complementary meta-analytic perspectives, highlighting their assumptions, strengths, and limitations.

Standardised mean differences, odds ratios, and Fisher's z-transformed correlations can all be analysed in multilevel meta-regression and heterogeneity models, including moderator-dependent residual or study-level heterogeneity when the data support it. However, these effect measures do not themselves provide the same direct contrast in within-group variability. A mean-versus-variability workflow therefore needs a separate dispersion-focused effect size, such as lnVR or lnCVR, when suitable group-level data are available. In particular, the within-group SD in a standardised mean difference standardises the mean contrast; it does not directly estimate a between-group variability contrast.

For clarity, the four meta-analytic models illustrated in this tutorial are:

-   Model 1 (mean-effect meta-analysis): synthesis of mean effects using the log response ratio (lnRR)

-   Model 2 (meta-analysis of variability): synthesis of within-population variability using lnVR and lnCVR

-   Model 3 (location-scale meta-regression of mean effects): location-scale meta-regression applied to lnRR to examine how moderator(s) influence both the mean and its heterogeneity

-   Model 4 (location-scale meta-regression of variability): location-scale meta-regression applied to lnVR/lnCVR to examine how moderator(s) influence both within-population variability and the heterogeneity of variability


# Introduction

This online tutorial walks you through each step of meta-analysis and location-scale meta-regression using three `R` packages: `metafor`, `brms`, and `glmmTMB`.
With one real dataset, we show how to analyse mean effects, within-population variability, and context-dependent heterogeneity, from data preparation through model interpretation.

Detailed methodological explanations and additional examples are beyond the scope of this tutorial - readers interested in further details may find the following website and relevant papers helpful:

1.  **Meta-analysis**\
    Online tutorial: <https://itchyshin.github.io/Meta-analysis_tutorial/>\
    Related paper:\
    Nakagawa S, Yang Y, Macartney EL, Spake R, Lagisz M.
    (2023).
    Quantitative evidence synthesis: a practical guide on meta-analysis, meta-regression, and publication bias tests for environmental sciences.
    *Environmental Evidence.* 12:8.
    <https://doi.org/10.1186/s13750-023-00301-6>

2.  **Location–scale meta-analysis models**\
    Online tutorial: <https://itchyshin.github.io/location-scale_meta-analysis/>\
    Related paper:\
    Nakagawa S, Mizuno A, Morrison K, Ricolfi L, Williams C, Drobniak SM, Lagisz M, Yang Y.
    (2025).
    Location‐Scale Meta‐Analysis and Meta‐Regression as a Tool to Capture Large‐Scale Changes in Biological and Methodological Heterogeneity: A Spotlight on Heteroscedasticity.
    *Global Change Biology.* 31:e70204.
    <https://doi.org/10.1111/gcb.70204>

3.  **Location–scale models**\
    Online tutorial: <https://ayumi-495.github.io/Eco_location-scale_model/>\
    Related paper:\
    Nakagawa S, Ortega S, Gazzea E, Lagisz M, Lenz A, Lundgren E, Mizuno A.
    (2025).
    Location–scale models in ecology and evolution: Heteroscedasticity in continuous, count and proportion data.
    *Methods in Ecology and Evolution.* <https://doi.org/10.1111/2041-210x.70203>

::: callout-note
Outputs for the scale component in scale parts are reported on the log scale.
Guidance on how to interpret these parameters, including back-transformation and interpretation is provided in the third online tutorial.
:::

## Setup

### Load packages

First, we need to load necessary packages.

```{r}
# needed packages
pacman::p_load("brms", "glmmTMB", "metafor", "tidyverse", 
               "tidybayes", "orchaRd", "here")

# Cached fitted objects are deliberately separate from the source document.
# From revision/, regenerate them with:
# Rscript R/regenerate_tutorial_artifacts.R
# Rscript R/study_level_heterogeneity.R --fit-bayesian
read_cached_artifact <- function(name) {
  path <- here::here("Rdata", name)
  if (!file.exists(path)) {
    stop(
      "Missing cached artifact: ", path, "\n",
      "Run the documented regeneration command from revision/ before rendering.",
      call. = FALSE
    )
  }
  readRDS(path)
}
```

::: callout-note
On Windows, `brm()` compiles Stan code and requires an [Rtools version compatible with your installed R version](https://cran.r-project.org/bin/windows/Rtools/).
:::

::: callout-important
Until recently `equalto()`, which is required to conduct meta-analyses in `glmmTMB`, was not implemented.
If you want to run meta-analytic models (including meta-regression), please ensure that you have installed the latest version of the package.

Also, we have created a custom function for visualising the outputs of meta-analysis models fitted in `brms` If you are interested, please expand the folded code below to view it.
:::

```{r}
#| code-fold: true

library(patchwork)
get_variables_dynamic <- function(model, patterns) {
  variables <- get_variables(model)
  idx <- unique(unlist(lapply(patterns, function(p) grep(p, variables))))
  if (length(idx) == 0) return(character(0))
  variables[idx]
}

# preparation
rename_vars <- function(variable) {
  # fixed effects
  variable <- gsub("b_Intercept", "b_l_intercept", variable)
  variable <- gsub("b_sigma_Intercept", "b_s_intercept", variable)  
  variable <- gsub("b_Org.fertilizer.typeplant", "b_l_contrast", variable)
  variable <- gsub("b_sigma_Org.fertilizer.typeplant", "b_s_contrast", variable)            
  # random effects
  variable <- gsub("sd_study_ID__Intercept", "sd_study_ID", variable)  
  variable <- gsub("sigma", "sd_effect_ID", variable)  
  return(variable)
}

# Function to visualise fixed effects
visualize_fixed_effects <- function(model) {
  fixed_effect_vars <- get_variables_dynamic(model, "^b_")
  if (length(fixed_effect_vars) == 0) {
    message("No fixed effects found")
    return(NULL)
  }
  
  tryCatch({
    fixed_effects_samples <- model %>%
      spread_draws(!!!syms(fixed_effect_vars)) %>%
      pivot_longer(cols = all_of(fixed_effect_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(
        .variable = factor(
          rename_vars(.variable),
          levels = rev(c("b_l_intercept", "b_l_contrast",
                         "b_s_intercept", "b_s_contrast"))
        )
      )
    
    ggplot(fixed_effects_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        point_interval = "mean_qi", 
        fill = "lightcyan3", 
        color = "lightcyan4"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Fixed effects", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_fixed_effects: ", e$message)
    return(NULL)
  })
}

# Function to visualize random effects
visualize_random_effects <- function(model) {
  random_effect_vars <- get_variables_dynamic(model, c("^sd_study", "^sigma"))
  #random_effect_vars <- random_effect_vars[random_effect_vars != "sd_es_ID__Intercept"]
  if (length(random_effect_vars) == 0) {
    message("No random effects found")
    return(NULL)
  }
  
  tryCatch({
    random_effects_samples <- model %>%
      spread_draws(!!!syms(random_effect_vars)) %>%
      pivot_longer(cols = all_of(random_effect_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(.variable = rename_vars(.variable)) #%>%
    #mutate(.value = .value)  # leave SD as it is
    
    ggplot(random_effects_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        point_interval = "mean_qi", 
        fill = "olivedrab3", 
        color = "olivedrab4"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Random effects (SD)", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_random_effects: ", e$message)
    return(NULL)
  })
}

# Function to visualize random effects
visualize_random_effects2 <- function(model) {
  random_effect_vars <- get_variables_dynamic(model, c("^sd_study"))
  #random_effect_vars <- random_effect_vars[random_effect_vars != "sd_es_ID__Intercept"]
  if (length(random_effect_vars) == 0) {
    message("No random effects found")
    return(NULL)
  }
  
  tryCatch({
    random_effects_samples <- model %>%
      spread_draws(!!!syms(random_effect_vars)) %>%
      pivot_longer(cols = all_of(random_effect_vars), names_to = ".variable", values_to = ".value") %>%
      mutate(.variable = rename_vars(.variable)) #%>%
    #mutate(.value = .value)  # leave SD as it is
    
    ggplot(random_effects_samples, aes(x = .value, y = .variable)) +
      stat_halfeye(
        normalize = "xy", 
        point_interval = "mean_qi", 
        fill = "olivedrab3", 
        color = "olivedrab4"
      ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#005") +
      labs(y = "Random effects (SD)", x = "Posterior values") +
      theme_classic()
  }, error = function(e) {
    message("Error in visualize_random_effects: ", e$message)
    return(NULL)
  })
}

```

All the outputs from the code chunks are saved in the [Google drive folder](https://drive.google.com/drive/folders/1gLn52s5hL_4JI0Z0-t-nmIjeur-_0v4n?usp=sharing) to speed up the rendering process of this tutorial. You can download all files and see these outputs in your R environment.
If you want to run all the code chunks by yourself, please set `eval: true` for those code chunks with `eval: false`.

### Data preparation

We use a dataset from Ponisio et al. (2015), which synthesises experimental studies comparing organic and conventional crop yields across a wide range of crop types, management practices, and geographic regions.
The full meta-dataset comprises over 1,000 organic–conventional yield comparisons drawn from more than 100 primary studies; detailed metadata, inclusion criteria, and data descriptions are provided in the original publication and its associated supplementary materials.\
A subset of the data is used here to facilitate computation and clarity of presentation while preserving the essential structure of the dataset.

Download [`ponisio2014dataset.csv`](https://raw.githubusercontent.com/Ayumi-495/one-dataset-four-meta-analyses/main/data/ponisio2014dataset.csv) directly from this repository.
Create a folder named `data` in the root of your R project, save the downloaded file there as `ponisio2014dataset.csv`, and open the project before running the tutorial.
The code below then loads the file without modification using `here::here("data", "ponisio2014dataset.csv")`.

```{r}

datfull <- read.csv(here::here("data", "ponisio2014dataset.csv"), header = TRUE)
# dim(datfull)
# [1] 1071   70 -> 1071 effect sizes and 70 columns
# head(datfull)　# check column names
# create ID using Author + Year + Journal - gives unique identifier for each study
datfull$study_ID <- as.numeric(as.factor(paste(datfull$Author, datfull$Year, datfull$Journal, sep = "_")))
datfull$effect_size_ID <- 1:nrow(datfull)

# filter to the analysis subset
# Org.fertilizer.type = either plant or animal and also Crop.type = cereals
dat <- datfull %>%
 dplyr::filter(Org.fertilizer.type %in% c("plant", "animal")) %>%
 dplyr::filter(Crop.type == "cereals") %>%
  droplevels()

# dim(dat) # 318 effect sizes

```

### Effect size calculation

Original study calculated and used log response ratio (lnRR) and its variance.
We will also calculate log variability ratio (lnVR) and log coefficient of variation ratio (lnCVR), and their variances, respectively.
`metafor` has a function `escalc` to calculate various effect sizes including lnVR and lnCVR.

```{r}

dat <- escalc(measure = "VR",
                      m1i = Mean.Org,
                      sd1i = SD.Org,
                      n1i = N.Org,
                      m2i = Mean.conv,
                      sd2i = SD.conv,
                      n2i = N.Conv,
                      data = dat,
                      var.names = c("lnVR", "var.lnVR")
              )

dat <- escalc(measure = "CVR",
              m1i = Mean.Org,
              sd1i = SD.Org,
              n1i = N.Org,
              m2i = Mean.conv,
              sd2i = SD.conv,
              n2i = N.Conv,
              data = dat,
              var.names = c("lnCVR", "var.lnCVR"),
              correct = TRUE
              )

```

The `correct = TRUE` argument explicitly applies the second-order Taylor bias correction for lnCVR. This correction became the default in `metafor` 5.0-1. Specifying this argument explicitly makes the bias-correction choice reproducible across `metafor` versions in which the default differs.

# Model 1: Meta-analysis of mean

We will first fit multilevel meta-analysis and meta-regression models to estimate the overall mean effect size using lnRR effect size.
We will model effect sizes as the response variable, with random intercepts specified for study identity and effect-size identity to account for the hierarchical structure of the data and non-independence among observations.

## metafor

:::: panel-tabset
### meta-analysis (overall effect)

When we want to run the multilevel meta-analysis using `metafor`, we can use the `rma.mv` function.
Here, we specify `lnRR` as the response variable and `var.lnRR` as the sampling variance.
Random effects are specified using the `random` argument, where we include random intercepts for both `study_ID` and `effect_size_ID`.

```{r}
#| eval: false

mod_lnRR <- rma.mv(lnRR,
                    var.lnRR,
                    mod = ~ 1,
                    random = list(~ 1 | study_ID,
                                  ~ 1 | effect_size_ID),
                    data = dat,
                    test = "t",
                    method = "REML"
                   )
summary(mod_lnRR)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -109.8272   219.6545   225.6545   236.9312   225.7311   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0736  0.2712     36     no        study_ID 
# sigma^2.2  0.0830  0.2881    318     no  effect_size_ID 
# 
# Test for Heterogeneity:
# Q(df = 317) = 15841.3280, p-val < .0001
# 
# Model Results:
# 
# estimate      se     tval   df    pval    ci.lb    ci.ub      
#  -0.3204  0.0534  -5.9951  317  <.0001  -0.4255  -0.2152  *** 
# 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

i2_ml(mod_lnRR)
         # I2_Total       I2_study_ID I2_effect_size_ID 
         # 98.11404          46.09489          52.01915 

```

```{r}
#| echo: false
#| fig-align: center

orchard_ma_lnRR <- read_cached_artifact("orchard_ma_lnRR.rds")
orchard_ma_lnRR

```

The estimated average effect size was negative ($\beta_0 = -0.32$, 95% CI = [-0.43, -0.22]), indicating that organic farming reduced crop yield by approximately 27% compared to conventional farming ($1 - \exp(-0.32) = 1 - 0.73$).
There was also high heterogeneity among effect sizes ($I^2_{\text{Total}} = 98.11\%$), with contributions from both the study level and the effect size level ($I^2_{\text{between}} = 46.09\%$, $I^2_{\text{within}} = 52.02\%$).
This indicates that effect sizes vary widely across studies as well as within studies.

### meta-regression

Next, we fit a simple meta-regression to examine the type of fertiliser (animal- vs plant-based) which may explain variation in effect sizes.

```{r}
#| eval: false

mod_lnRR_mr <- rma.mv(lnRR,
                    var.lnRR,
                    mods = ~ 1 + Org.fertilizer.type,
                    random = list(~ 1 | study_ID,
                                  ~ 1 | effect_size_ID),
                    data = dat,
                    test = "t",
                    method = "REML"
                    )
summary(mod_lnRR_mr)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -106.8735   213.7470   221.7470   236.7699   221.8756   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0614  0.2477     36     no        study_ID 
# sigma^2.2  0.0832  0.2885    318     no  effect_size_ID 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 13892.9906, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 5.7785, p-val = 0.0168
# 
# Model Results:
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -0.2663  0.0552  -4.8269  316  <.0001  -0.3749  -0.1578  *** 
# Org.fertilizer.typeplant   -0.1407  0.0585  -2.4038  316  0.0168  -0.2558  -0.0255    * 

r2_ml(mod_lnRR_mr)
# R2_marginal R2_conditional 
#   0.0323952      0.4429578 

```

```{r}
#| echo: false
#| fig-align: center

mod_lnRR_mr <- read_cached_artifact("mod_lnRR_mr_metafor.rds")

orchard_plot(mod_lnRR_mr, 
             mod = "Org.fertilizer.type",
             xlab = "lnRR", 
             group = "study_ID",
             trunk.size = 0.6,
             branch.size = 5,
             mod.order = c("Plant", "Animal")) +
  scale_colour_manual(values = rep("grey20",8)) +
  scale_fill_manual(values = c("#CC6677", "#CC6677"))

```

The intercept represents the estimated mean lnRR for studies using animal-based fertilisers.
The coefficient for plant-based fertilisers quantifies the difference in lnRR between plant- and animal-based fertilisers.

In the model summary, the estimated coefficient for plant-based fertilisers was negative and statistically significant, indicating that yields under plant-based organic fertilisers were, on average, lower than those under animal-based organic fertilisers.
The moderator explained approximately 3.2% of the total variance in effect sizes ($R^2_{\text{marginal}} = 0.032$), indicating that differences between fertiliser types account for only a small proportion of the overall heterogeneity.

This pattern is illustrated in the orchard plot, which shows individual effect size estimates (lnRR) for plant- and animal-based fertilisers, with point sizes scaled by precision (1/SE).
The black squares represent the model-predicted mean effects for each fertiliser type, with horizontal lines indicating their associated uncertainty intervals.
Although the mean lnRR differs between fertiliser types, the substantial overlap and wide spread of individual effect sizes highlight that most of the variability remains unexplained by fertiliser type alone.

::: callout-note
### $I^2$ vs $R^2$

$I^2$ and $R^2$ are both measures used in meta-analysis, but they quantify different aspects of variability in effect sizes.

$I^2$ quantifies the proportion of the total observed variance in effect size estimates that is attributable to modelled heterogeneity rather than sampling variance.
Its interpretation depends on the specified model structure and on which sources of heterogeneity are included, particularly in multilevel or location–scale meta-analytic models.

By contrast, $R^2$ quantifies the proportion of variance in effect sizes that is explained by moderators in a meta-regression framework, conditional on the fitted model.
$R^2_{\text{marginal}}$ reflects the extent to which the included moderators explain heterogeneity, whereas $R^2_{\text{conditional}}$ represents the proportion of variance explained by the full model, including both fixed effects and random effects, under the specified variance structure.
:::
::::

## brms

As we mentioned in the main text, we can use the `brms` package to fit the same multilevel meta-analysis model.
There are different ways to run these models, but in this example, we use the `gr()` function to set the sampling variance-covariance matrix.
This approach is less common, but it can be much faster than usual (see more detailed information: [link](https://itchyshin.github.io/location-scale_meta-analysis/#brms)).

If you are not familiar with `brms`, we highly recommend checking out the instruction of [brms](https://paulbuerkner.com/brms/index.html) first.

::: panel-tabset
### meta-analysis (overall effect)

In the model below, sampling error is explicitly represented using a group-level term with a known variance–covariance structure.
The term `(1 | gr(effect_size_ID, cov = vcv))` specifies a random effect whose covariance matrix is fixed to the diagonal matrix of sampling variances for lnRR.
This formulation is mathematically equivalent to supplying a sampling variance matrix in `metafor::rma.mv()`, but implemented here using `brms`.

Because sampling variances are assumed to be known in meta-analysis, the standard deviation of this group-level effect is fixed to *1*.
As a result, the magnitude of sampling error is fully determined by the supplied variance–covariance matrix and is not estimated from the data.

```{r}
#| eval: false

vcv <- diag(dat$var.lnRR)

rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_ma1 <- bf(lnRR ~ 1 +
                 (1|study_ID) + # this is u_l (the between-study effect)
                 (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_ma1 <- default_prior(form_ma1, 
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_ma1$prior[3] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior_ma1

# fitting model
fit_ma1 <- brm(
  formula = form_ma1,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_ma1,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_ma1)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: lnRR ~ 1 + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.28      0.04     0.21     0.38 1.00     1839     3325
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept    -0.32      0.06    -0.43    -0.21 1.00     1842     2814
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.29      0.01     0.26     0.32 1.00     7189     4944
# 
# Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
# and Tail_ESS are effective sample size measures, and Rhat is the potential
# scale reduction factor on split chains (at convergence, Rhat = 1).

```

```{r}
#| echo: false
#| fig-align: center

fit_ma1 <- read_cached_artifact("fit_ma1.rds")
p1 <- visualize_fixed_effects(fit_ma1)
p2 <- visualize_random_effects(fit_ma1)
p1 + p2

```

The estimated overall mean and between-study standard deviation are comparable to the study-level variance component obtained from the `metafor` model.
In contrast, the residual parameter `sigma` represents unexplained heterogeneity at the effect-size level beyond sampling error, corresponding to the effect-size-level variance component in `rma.mv()`.

One important point concerns how random effects are reported.
In `brms`, random-effect estimates are presented as *standard deviations (SD)*, whereas `metafor` reports variances.
Therefore, to compare estimates between the two packages, the standard deviations reported by `brms` must be squared to obtain the corresponding variance estimates.

Finally, small numerical differences between estimates obtained from `brms` and `metafor` are expected, even when the underlying model structures are closely aligned.
This is because `brms` fits models using Bayesian inference, which relies on probabilistic sampling to approximate posterior distributions, whereas `metafor` uses frequentist estimation.
As a result, point estimates may differ slightly due to Monte Carlo error and differences in estimation frameworks, rather than reflecting substantive discrepancies between the models.

### meta-regression

Let's now fit a meta-regression model including fertiliser type as a moderator - the outputs are very similar to those from `metafor`.

```{r}
#| eval: false

vcv <- diag(dat$var.lnRR)
rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_mr1 <- bf(lnRR ~ 1 + Org.fertilizer.type +
                  (1|study_ID) + # this is u_l (the between-study effect)
                  (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_mr1 <- default_prior(form_mr1,
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_mr1$prior[5] = "constant(1)" # meta-analysis assumes

# fitting model
fit_mr1 <- brm(
  formula = form_mr1,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_mr1,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_mr1)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.26      0.04     0.19     0.35 1.00     2384     4173
# 
# Regression Coefficients:
#                          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept                   -0.27      0.06    -0.38    -0.15 1.00     2787     3503
# Org.fertilizer.typeplant    -0.14      0.06    -0.26    -0.02 1.00     4423     4398
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.29      0.01     0.26     0.32 1.00     6530     4799
# 
# Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
# and Tail_ESS are effective sample size measures, and Rhat is the potential
# scale reduction factor on split chains (at convergence, Rhat = 1).

```

```{r}
#| echo: false

fit_mr1 <- read_cached_artifact("fit_mr1.rds")

```

You may wonder why we specify `~1 + Org.fertilizer.type` rather than simply `~ Org.fertilizer.type`.
In fact, there is no difference between these two specifications.
In R model formulas, an intercept term is included by default, so both formulas fit a model with an intercept.

We write 1 explicitly here for clarity.
Doing so makes it clear that the model includes an overall intercept, which represents the baseline mean effect size for the reference level of the moderator.
This is particularly helpful in a meta-analytic context, where the intercept has a clear interpretation as the overall mean effect.
For this reason, we explicitly include 1 in all model formulas throughout this tutorial.
This convention makes the role of the intercept transparent and helps avoid confusion when comparing different model specifications.
:::

## glmmTMB

`glmmTMB` is a popular package for fitting generalised linear mixed models.
Unlike dedicated meta-analytic packages such as `metafor`, it does not allow to easily account for known sampling variances of effect sizes - so when we implement a meta-analysis in `glmmTMB`, we need explicitly specify the sampling error structure of the effect sizes.

In practice, this require three steps:

*1. Treating each effect size as a unique random-effect level*

```{r}

dat$effect_size_ID <- factor(dat$effect_size_ID)
```

In `glmmTMB`, grouping variables for random effects must be factors.
By converting `effect_size_ID` to a factor, we ensure that each effect size is treated as a distinct level in the random-effects structure.
This step is required for the model to correctly associate each observation with its corresponding sampling variance.

*2. Specifying the random effect independent grouping structure*

```{r}

dat$g <- 1
```

*3. Constructing a variance-covariance matrix (VCV) that contains the known sampling variances of each effect size*

```{r}

vcv <- diag(dat$var.lnRR)

row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID
```

We create a variance–covariance matrix whose diagonal elements are the known sampling variances of the effect sizes (`var.lnRR`).
Off-diagonal elements are set to zero, reflecting the assumption that effect sizes are independent.
This mirrors the standard assumption made in most meta-analyses.

The `equalto()` term in `glmmTMB` requires a grouping factor.
The dummy variable `g` serves this purpose, which we set to a single level (one group), i.e. all observations are assigned to the same group. If the grouping factor had more than one level, each group would have a separate independent random-effect vector with the same variance-covariance parameters. For meta-analysis, this grouping factor should always have one level.

For more details, you can also refer to this vignette: <https://cran.r-project.org/web/packages/glmmTMB/vignettes/covstruct.html>

And the following supplementary material for fitting meta-analysis with `glmmTMB`:
<https://coraliewilliams.github.io/equalto_sim_study/webpage.html>.

::: panel-tabset
### meta-analysis (overall effect)

```{r}
#| eval: false

glmm_lnRR_ma <- glmmTMB(lnRR ~ 1 +
                        (1 | study_ID) + 
                         equalto(0 + effect_size_ID|g, vcv), #  incorporates the known sampling variances of each effect size
                         data = dat,
                        REML = TRUE
                        )
summary(glmm_lnRR_ma)
#  Family: gaussian  ( identity )
# Formula:          lnRR ~ 1 + (1 | study_ID) + equalto(0 + effect_size_ID | g, vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid
#     231.4     242.7    -112.7     225.4       315
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr
#  study_ID (Intercept)      0.0735694 0.27124
#  g        effect_size_ID33 0.0047213 0.06871
#           effect_size_ID34 0.0426425 0.20650  0.00
#           effect_size_ID35 0.0698313 0.26426  0.00 0.00
#           effect_size_ID36 0.0798567 0.28259  0.00 0.00 0.00
#           effect_size_ID37 0.0837435 0.28938  0.00 0.00 0.00 0.00
#           effect_size_ID39 0.0095870 0.09791  0.00 0.00 0.00 0.00 0.00
#           effect_size_ID40 0.0091520 0.09567  0.00 0.00 0.00 0.00 0.00 0.00
#           effect_size_ID42 0.0363323 0.19061  0.00 0.00 0.00 0.00 0.00 0.00 0.00
#           effect_size_ID44 0.0331452 0.18206  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00
#           effect_size_ID45 0.0012338 0.03513  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.0830247 0.28814
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.083
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)
# (Intercept) -0.32039    0.05358   -5.98 2.23e-09 ***

```

In short, you should interpret the output as follows:

-   Overall effect estimate: `(Intercept)`

-   Between-study variance estimate: `study_ID (Intercept)` variance

-   Effect-size–level (sampling) variance: `Residual` / `sigma^2`

All estimate is identical to `metafor` output, and also close to `brms` one.

NOTE:\
The grouping factor `g` is a technical dummy variable required by `glmmTMB` when using `equalto()` to impose a known variance-covariance structure.
All observations belong to a single level of `g`, and the variance structure across effect sizes is entirely defined by the supplied VCV matrix.

As a result, the multiple variance components listed under `g` in the model output correspond to the known sampling variances supplied via the VCV matrix rather than free parameters estimated by the model.
A useful visual check is to confirm that each `effect_size_ID` is associated with an appropriate variance value that matches its sampling variance.
The output also makes explicit that independence among effect sizes is assumed, as indicated by the zero entries in the `Corr` columns, which provides an additional quick diagnostic check (although substantive interest typically lies in the study-level variance and the residual variance).

::: callout-important
`0 +` is required when using a structured covariance matrix in `glmmTMB` because it removes the implicit intercept and ensures a one-to-one correspondence between the random-effect coefficients and the rows and columns of the supplied covariance matrix; otherwise, the covariance structure is mis-specified (see - <https://cran.r-project.org/web/packages/glmmTMB/vignettes/covstruct.html#construction-of-structured-covariance-matrices>).
:::

### meta-regression

When you want to run the meta-regression, you will simply add the moderators in the model:

```{r}
#| eval: false

glmm_lnRR_mr <- glmmTMB(lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + 
                      equalto(0 + effect_size_ID|g, vcv),
                  data = dat,
                  REML = TRUE
                  )
summary(glmm_lnRR_mr)
#  Family: gaussian  ( identity )
# Formula:          lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +  
#     effect_size_ID | g, vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     231.9     246.9    -111.9     223.9       314 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr                                            
#  study_ID (Intercept)      0.0613541 0.24770                                                  
#  g        effect_size_ID33 0.0047213 0.06871                                                  
#           effect_size_ID34 0.0426425 0.20650  0.00                                            
#           effect_size_ID35 0.0698313 0.26426  0.00 0.00                                       
#           effect_size_ID36 0.0798567 0.28259  0.00 0.00 0.00                                  
#           effect_size_ID37 0.0837435 0.28938  0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.0095870 0.09791  0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.0091520 0.09567  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.0363323 0.19061  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.0331452 0.18206  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.0012338 0.03513  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.0832439 0.28852                                                  
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.0832 
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -0.26633    0.05522  -4.823 1.41e-06 ***
# Org.fertilizer.typeplant -0.14069    0.06008  -2.342   0.0192 * 

```

Great - the estimates match those from `metafor` and are also very close to the results from `brms` :)
:::

# Model 2: Meta-analysis of variability

In this section we analyse variability effects using **lnCVR** and **lnVR**.

-   **lnCVR** quantifies the log ratio of coefficients of variation (CVs), that is, changes in **relative variability** after accounting for mean differences.

-   **lnVR** quantifies the log ratio of standard deviations (SDs), that is, changes in **absolute variability** between groups.

## metafor

::: panel-tabset
### meta-analysis (overall effect)

#### lnCVR

```{r}
#| eval: false

mod_lnCVR <- rma.mv(lnCVR,
                   var.lnCVR,
                   random = list(~ 1 | study_ID,
                                 ~ 1 | effect_size_ID),
                   data = dat,
                   test = "t",
                   method = "REML"
                   )
summary(mod_lnCVR)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -437.7598   875.5196   881.5196   892.7963   881.5963   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0583  0.2416     36     no        study_ID 
# sigma^2.2  0.2900  0.5385    318     no  effect_size_ID 
# 
# Test for Heterogeneity:
# Q(df = 317) = 795.4614, p-val < .0001
# 
# Model Results:
# 
# estimate      se    tval   df    pval   ci.lb   ci.ub      
#   0.3397  0.0702  4.8421  317  <.0001  0.2017  0.4778  *** 
# 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

i2_ml(mod_lnCVR)
# I2_Total       I2_study_ID I2_effect_size_ID 
# 55.863504          9.357345         46.506159

```

```{r}
#| echo: false
#| fig-align: center

orchard_ma_lnCVR <- read_cached_artifact("orchard_ma_lnCVR.rds")
orchard_ma_lnCVR

```

The intercept from the multilevel meta-analysis represents the overall mean lnCVR across all comparisons.
A positive estimate indicates that the treatment group has a larger CV than the control group, indicating greater mean-corrected within-group variability after accounting for potential differences in the mean values.

Here, the estimated lnCVR is $\beta_0 = 0.34$ (95% CI = [0.20, 0.48]), suggesting that organic farming is associated with approximately 41% higher relative variability in crop yields compared to conventional farming ($\exp(0.34) - 1$).

$I^2$ values indicate that a substantial portion of the total heterogeneity ($I^2_{\text{Total}} = 55.86\%$) arises from variability among effect sizes within studies ($I^2_{\text{within}} = 46.51\%$), while a smaller share is due to variability between studies ($I^2_{\text{between}} = 9.36\%$).

#### lnVR

```{r}
#| eval: false

mod_lnVR <- rma.mv(lnVR,
                   var.lnVR,
                   random = list(~ 1 | study_ID,
                                 ~ 1 | effect_size_ID),
                   data = dat,
                   test = "t",
                   method = "REML"
                   )
summary(mod_lnVR)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -423.3326   846.6651   852.6651   863.9418   852.7418   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0640  0.2530     36     no        study_ID 
# sigma^2.2  0.1886  0.4342    318     no  effect_size_ID 
# 
# Test for Heterogeneity:
# Q(df = 317) = 658.8664, p-val < .0001
# 
# Model Results:
# 
# estimate      se    tval   df    pval    ci.lb   ci.ub    
#   0.0198  0.0676  0.2924  317  0.7702  -0.1133  0.1528    
# 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

i2_ml(mod_lnVR)
# I2_Total       I2_study_ID I2_effect_size_ID 
# 50.62070          12.82949          37.79121 

```

```{r}
#| echo: false
#| fig-align: center

orchard_ma_lnVR <- read_cached_artifact("orchard_ma_lnVR.rds")
orchard_ma_lnVR

```

The overall lnVR estimate is close to zero and the 95% CI overlaps zero ($\beta_0 = 0.02$, 95% CI = [-0.11, 0.15]), indicating little evidence for a systematic difference in SD between the two groups.
On the original scale, this corresponds to an SD ratio close to one, suggesting that the magnitude of within-group variation is broadly similar between the two comparison groups.

Despite the absence of a clear average lnVR effect, heterogeneity is substantial.
The test for heterogeneity is highly significant, and approximately half of the observed variability in lnVR cannot be attributed to sampling error alone ($I^2 \approx 50.6\%$).
Only a smaller proportion of this heterogeneity is attributable to between-study differences ($I^2 \approx 12.8\%$), whereas a larger share arises at the effect-size level within studies ($I^2 \approx 37.8\%$).
This pattern indicates considerable context dependence or within-study variation in absolute variability effects, even though the average effect across all comparisons is close to zero.

When considered alongside lnCVR, a clearer picture emerges.
While lnVR shows little evidence for a consistent difference in absolute variability, lnCVR indicates a positive average effect, suggesting higher relative variability (CV) in the treatment group.
This divergence arises because lnCVR explicitly standardises variability by the mean, whereas lnVR does not.
When group means differ, absolute variability may remain similar ($\mathrm{lnVR} \approx 0$) while relative variability increases ($\mathrm{lnCVR} > 0$).

In practical terms, this pattern implies that outcomes may scale with the mean.
Although the absolute spread of values is comparable across groups, the spread is larger relative to the expected outcome under the treatment condition.
Consequently, differences in variability are more apparent on the mean-standardised scale than on the absolute scale.

### meta-regression

#### lnCVR

```{r}
#| eval: false

mod_lnCVR_mr <- rma.mv(lnCVR,
                    var.lnCVR,
                    mods = ~ Org.fertilizer.type,
                    random = list(~ 1 | study_ID,
                                  ~ 1 | effect_size_ID),
                    data = dat,
                    test = "t",
                    method = "REML"
                    )
summary(mod_lnCVR_mr)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -436.6658   873.3317   881.3317   896.3546   881.4603   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0583  0.2414     36     no        study_ID 
# sigma^2.2  0.2925  0.5408    318     no  effect_size_ID 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 795.3961, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 0.3085, p-val = 0.5790
# 
# Model Results:
# 
#                           estimate      se    tval   df    pval    ci.lb   ci.ub      
# intrcpt                     0.3117  0.0866  3.6001  316  0.0004   0.1413  0.4820  *** 
# Org.fertilizer.typeplant    0.0645  0.1161  0.5554  316  0.5790  -0.1639  0.2928      

r2_ml(mod_lnCVR_mr)
   # R2_marginal R2_conditional 
   #  0.002888785    0.168597379 

```

```{r}
#| echo: false
#| fig-align: center

mod_lnCVR_mr <- read_cached_artifact("mod_lnCVR_mr_metafor.rds")
orchard_plot(mod_lnCVR_mr, 
             mod = "Org.fertilizer.type",
             xlab = "lnCVR", 
             group = "study_ID",
             trunk.size = 0.6,
             branch.size = 5,
             mod.order = c("Plant", "Animal")) +
  scale_colour_manual(values = rep("grey20",8)) +
  scale_fill_manual(values = c("#117733", "#117733"))

```

#### lnVR

```{r}
#| eval: false

mod_lnVR_mr <- rma.mv(lnVR,
                      var.lnVR,
                      mods = ~ Org.fertilizer.type,
                      random = list(~ 1 | study_ID,
                                    ~ 1 | effect_size_ID),
                      data = dat,
                      test = "t",
                      method = "REML"
                      )
summary(mod_lnVR_mr)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -421.3055   842.6109   850.6109   865.6339   850.7396   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0560  0.2367     36     no        study_ID 
# sigma^2.2  0.1896  0.4355    318     no  effect_size_ID 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 651.1955, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 2.4353, p-val = 0.1196
# 
# Model Results:
# 
#                           estimate      se     tval   df    pval    ci.lb   ci.ub    
# intrcpt                     0.0876  0.0791   1.1075  316  0.2689  -0.0680  0.2433    
# Org.fertilizer.typeplant   -0.1644  0.1054  -1.5605  316  0.1196  -0.3717  0.0429    

r2_ml(mod_lnVR_mr)
# R2_marginal R2_conditional 
#  0.02621268     0.24827290 

```

```{r}
#| echo: false
#| fig-align: center

mod_lnVR_mr <- read_cached_artifact("mod_lnVR_mr_metafor.rds")
orchard_plot(mod_lnVR_mr, 
             mod = "Org.fertilizer.type",
             xlab = "lnVR", 
             group = "study_ID",
             trunk.size = 0.6,
             branch.size = 5,
             mod.order = c("Plant", "Animal")) +
  scale_colour_manual(values = rep("grey20",8)) +
  scale_fill_manual(values = c("#88CCEE", "#88CCEE"))

```

Including fertiliser type as a moderator provided little evidence that it systematically explains variability effects on either scale.
Here, the moderator coefficient represents the difference between plant-based and animal-based fertilisers (plant − animal).

For lnCVR, the estimated moderator effect was small and highly uncertain ($\beta_{\text{plant-animal}} = 0.064$, 95% CI = [-0.17, 0.29]), indicating that relative variability does not differ consistently between fertiliser categories.

Similarly, for lnVR, there was no clear support for fertiliser-type differences in absolute variability ($\beta_{\text{plant-animal}} = -0.164$, 95% CI = [-0.37, 0.04]).
Although the point estimate suggested a tendency toward lower lnVR for plant-based fertilisers relative to animal-based fertilisers, this effect was highly uncertain and not statistically supported.

In both models, substantial residual heterogeneity remained after accounting for fertiliser type, with large effect-size–level variance components (lnCVR: $\sigma^2 = 0.29$; lnVR: $\sigma^2 = 0.19$).
This indicates that neither relative nor absolute variability effects are well captured by this single moderator.
:::

## brms

When fitted with equivalent model structures, the `metafor` and `brms` implementations yield nearly identical estimates for both the overall effects and variance components.

::: panel-tabset
### meta-analysis (overall effect)

#### lnCVR

```{r}
#| eval: false

vcv <- diag(dat$var.lnCVR)

rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_ma2 <- bf(lnCVR ~ 1 +
                 (1|study_ID) + # this is u_l (the between-study effect)
                 (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_ma2 <- default_prior(form_ma2, 
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_ma2$prior[3] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1

# fitting model
fit_ma2 <- brm(
  formula = form_ma2,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_ma2,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )

summary(fit_ma2)
# Family: gaussian 
#   Links: mu = identity 
# Formula: lnCVR ~ 1 + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.26      0.08     0.12     0.43 1.00     1461     2280
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept     0.34      0.07     0.19     0.49 1.00     4969     4548
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.55      0.05     0.44     0.66 1.00      934     1852

```

```{r}
#| echo: false
#| fig-align: center

fit_ma2 <- read_cached_artifact("fit_ma2.rds")
p1 <- visualize_fixed_effects(fit_ma2) 
p2 <- visualize_random_effects(fit_ma2)
p1 + p2

```

#### lnVR

```{r}
#| eval: false

vcv <- diag(dat$var.lnVR)

rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_ma3 <- bf(lnVR ~ 1 +
                 (1|study_ID) + # this is u_l (the between-study effect)
                 (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_ma3 <- default_prior(form_ma3, 
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_ma3$prior[3] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1

# fitting model
fit_ma3 <- brm(
  formula = form_ma3,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_ma3,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_ma3)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: lnVR ~ 1 + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.27      0.07     0.15     0.42 1.00     1458     2196
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept     0.02      0.07    -0.12     0.17 1.00     2816     3765
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.44      0.06     0.32     0.56 1.01      469      856
```

```{r}
#| echo: false
#| fig-align: center

fit_ma3 <- read_cached_artifact("fit_ma3.rds")
p1 <- visualize_fixed_effects(fit_ma3) 
p2 <- visualize_random_effects(fit_ma3)
p1 + p2

```

### meta-regression

When specified with equivalent fixed and random effects and the same sampling variance structure, the `brms` and `metafor` meta-regression models yield nearly identical results.
Differences between outputs reflect estimation framework (Bayesian vs frequentist) rather than substantive differences in model inference.

#### lnCVR

```{r}
#| eval: false

vcv <- diag(dat$var.lnCVR)

rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_mr2 <- bf(lnCVR ~ 1 + Org.fertilizer.type +
                 (1|study_ID) + # this is u_l (the between-study effect)
                 (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_mr2 <- default_prior(form_mr2,
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_mr2$prior[5] = "constant(1)" # meta-analysis assumes

# fitting model
fit_mr2 <- brm(
  formula = form_mr2,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_mr2,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_mr2)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: lnCVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.26      0.08     0.12     0.44 1.00     1182     2170
# 
# Regression Coefficients:
#                          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept                    0.31      0.09     0.14     0.49 1.00     4348     4566
# Org.fertilizer.typeplant     0.06      0.12    -0.18     0.30 1.00     4027     4570
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.55      0.05     0.44     0.66 1.00     1026     1680

```

#### lnVR

```{r}
#| eval: false

vcv <- diag(dat$var.lnVR)

rownames(vcv) <- colnames(vcv) <- dat$effect_size_ID

form_mr3 <- bf(lnVR ~ 1 + Org.fertilizer.type +
                 (1|study_ID) + # this is u_l (the between-study effect)
                 (1|gr(effect_size_ID, cov = vcv)) # this is m (sampling error)
               )

# generate default priors
prior_mr3 <- default_prior(form_mr3,
                           data=dat, 
                           data2=list(vcv=vcv),
                           family=gaussian())
prior_mr3$prior[5] = "constant(1)" # meta-analysis assumes

# fitting model
fit_mr3 <- brm(
  formula = form_mr3,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_mr3,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_mr3)
# Family: gaussian 
#   Links: mu = identity 
# Formula: lnVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.24      0.07     0.11     0.39 1.00     1038     1472
# 
# Regression Coefficients:
#                          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept                    0.09      0.08    -0.06     0.25 1.00     3110     4251
# Org.fertilizer.typeplant    -0.17      0.11    -0.38     0.04 1.00     2781     3712
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.44      0.06     0.32     0.56 1.00      443      659

```
:::

## glmmTMB

We can also fit the same meta-analytic and meta-regression models using `glmmTMB`.
The outputs are very similar to those from `metafor` and `brms`.

First, convert the effect-size identifier and the technical grouping variable to factors. The response-specific sampling variance-covariance matrix is constructed in each model chunk below so that a matrix from a previous model cannot be reused accidentally.

```{r}
#| eval: false

dat$effect_size_ID <- factor(dat$effect_size_ID)
dat$g <- factor(1)

```

::: panel-tabset
### meta-analysis (overall effect)

```{r}
#| eval: false

vcv <- diag(dat$var.lnCVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnCVR <- glmmTMB(lnCVR ~ 1 + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      data = dat,
                      REML = TRUE
                      )
summary(glmm_lnCVR)
#  Family: gaussian  ( identity )
# Formula:          lnCVR ~ 1 + (1 | study_ID) + equalto(0 + effect_size_ID | g,      vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     887.3     898.6    -440.6     881.3       315 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance Std.Dev. Corr                                            
#  study_ID (Intercept)      0.05835  0.2416                                                   
#  g        effect_size_ID33 0.11583  0.3403                                                   
#           effect_size_ID34 0.29264  0.5410   0.00                                            
#           effect_size_ID35 0.31983  0.5655   0.00 0.00                                       
#           effect_size_ID36 0.32986  0.5743   0.00 0.00 0.00                                  
#           effect_size_ID37 0.33374  0.5777   0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.12070  0.3474   0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.12026  0.3468   0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.28633  0.5351   0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.28315  0.5321   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.25123  0.5012   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...      ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.28998  0.5385                                                   
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.290 
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)    
# (Intercept)  0.33973    0.07036   4.828 1.38e-06 ***

```

```{r}
#| eval: false

vcv <- diag(dat$var.lnVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnVR <- glmmTMB(lnVR ~ 1 + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      data = dat,
                      REML = TRUE
                     )
summary(glmm_lnVR)
#  Family: gaussian  ( identity )
# Formula:          lnVR ~ 1 + (1 | study_ID) + equalto(0 + effect_size_ID | g, vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     858.4     869.7    -426.2     852.4       315 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance Std.Dev. Corr                                            
#  study_ID (Intercept)      0.06401  0.2530                                                   
#  g        effect_size_ID33 0.11111  0.3333                                                   
#           effect_size_ID34 0.25000  0.5000   0.00                                            
#           effect_size_ID35 0.25000  0.5000   0.00 0.00                                       
#           effect_size_ID36 0.25000  0.5000   0.00 0.00 0.00                                  
#           effect_size_ID37 0.25000  0.5000   0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.11111  0.3333   0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.11111  0.3333   0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...      ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.18856  0.4342                                                   
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.189 
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)
# (Intercept)  0.01977    0.06787   0.291    0.771

```

### meta-regression

```{r}
#| eval: false

vcv <- diag(dat$var.lnCVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnCVR_mr <- glmmTMB(lnCVR ~ 1
                        + Org.fertilizer.type
                        + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      data = dat,
                      REML = TRUE
                      )
summary(glmm_lnCVR_mr)
#  Family: gaussian  ( identity )
# Formula:          lnCVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +      effect_size_ID | g, vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     891.4     906.5    -441.7     883.4       314 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance Std.Dev. Corr                                            
#  study_ID (Intercept)      0.05830  0.2415                                                   
#  g        effect_size_ID33 0.11583  0.3403                                                   
#           effect_size_ID34 0.29264  0.5410   0.00                                            
#           effect_size_ID35 0.31983  0.5655   0.00 0.00                                       
#           effect_size_ID36 0.32986  0.5743   0.00 0.00 0.00                                  
#           effect_size_ID37 0.33374  0.5777   0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.12070  0.3474   0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.12026  0.3468   0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.28633  0.5351   0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.28315  0.5321   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.25123  0.5012   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...      ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.29250  0.5408                                                   
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.292 
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)               0.31170    0.08728   3.571 0.000355 ***
# Org.fertilizer.typeplant  0.06446    0.11662   0.553 0.580457    

```

```{r}
#| eval: false

vcv <- diag(dat$var.lnVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnVR_mr <- glmmTMB(lnVR ~ 1
                        + Org.fertilizer.type
                        + (1 | study_ID) +
                        equalto(0 + effect_size_ID|g, vcv),
                      data = dat,
                      REML = TRUE
                      )
summary(glmm_lnVR_mr)
#  Family: gaussian  ( identity )
# Formula:          lnVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +      effect_size_ID | g, vcv)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     860.7     875.8    -426.4     852.7       314 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance Std.Dev. Corr                                            
#  study_ID (Intercept)      0.05602  0.2367                                                   
#  g        effect_size_ID33 0.11111  0.3333                                                   
#           effect_size_ID34 0.25000  0.5000   0.00                                            
#           effect_size_ID35 0.25000  0.5000   0.00 0.00                                       
#           effect_size_ID36 0.25000  0.5000   0.00 0.00 0.00                                  
#           effect_size_ID37 0.25000  0.5000   0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.11111  0.3333   0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.11111  0.3333   0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.25000  0.5000   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...      ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                  0.18965  0.4355                                                   
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 0.19 
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)
# (Intercept)               0.08762    0.07921   1.106    0.269
# Org.fertilizer.typeplant -0.16443    0.10620  -1.548    0.122

```
:::

# Models 3 and 4: Location-scale meta-regression models

We now extend the above meta-analyses to location–scale meta-regression models, which allow us to examine whether both the mean effect size and the variability of effect sizes depend on a moderator variable.
Here, we use `Org.fertilizer.type` (plant-based vs animal-based organic fertiliser) as a moderator.

-   The location component answers the question:\
    Does the mean effect size (i.e. lnRR, lnCVR, lnVR) differ between plant-based and animal-based organic fertilisers?

-   The scale component answers the question:\
    Does the residual heterogeneity (i.e. variability of effect sizes after accounting for random effects) differ between plant-based and animal-based organic fertilisers?

## metafor

At present, `metafor` does not support fully multilevel location–scale meta-analytic models (but it is currently under development - the random effects will soon be available to fit in scale!).
However, it is possible to fit a simpler location–scale model using the `scale` argument in `rma()`.
This approach allows the residual heterogeneity to vary as a function of moderators, but does not accommodate multilevel random-effects structures.

Below, we demonstrate two types of models:

A.  Simple location–scale meta-regression

B.  Multilevel heteroscedastic meta-regression

::: callout-note
It is best to treat the `metafor` models presented below as approximations to full multilevel location-scale meta-analytic models.
Although these models are extremely useful and sometimes sufficient in practice, ecological datasets often show more complex hierarchical structures, such as multiple effect sizes per study, evolutionary independence, or other sources of non-independence.
To explicitly consider such complexity, more flexible modelling frameworks are required.
In the following two sections, we introduce location-scale meta-analytic models fitted using `brms` and `glmmTMB`.
These approaches allow the specification of rich multilevel structures, and in particular, `brms` enables the fitting of fully Bayesian location-scale models with complex random effects and covariance structures (though if we want to run this type of model, we need a lot of data…)
:::

:::: panel-tabset
### A. Simple (no random effects on location and scale components)

The formulation using `rma()` provides a straightforward way to test whether variability in effect sizes depends on the moderator, but it treats all effect sizes as independent and does not include a multilevel structure.
`scale` argument allows us to specify a model for the residual variance, enabling us to test whether the variability of effect sizes differs across levels of the moderator.

#### lnRR

```{r}
#| eval: false

mod_lnRR_ls <- rma(yi = lnRR,
                   vi = var.lnRR,
                   mods = ~ Org.fertilizer.type,
                   scale = ~ Org.fertilizer.type,
                   data = dat,
                   test = "t",
                   method = "REML"
                   )
summary(mod_lnRR_ls)
# Location-Scale Model (k = 318; tau^2 estimator: REML)
# 
#    logLik   deviance        AIC        BIC       AICc   
# -161.2053   322.4106   330.4106   345.4336   330.5393   
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 13892.9906, p-val < .0001
# 
# Test of Location Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 67.0786, p-val < .0001
# 
# Model Results (Location):
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -0.1796  0.0290  -6.1935  316  <.0001  -0.2366  -0.1225  *** 
# Org.fertilizer.typeplant   -0.3666  0.0448  -8.1902  316  <.0001  -0.4547  -0.2785  *** 
# 
# Test of Scale Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 19.9223, p-val < .0001
# 
# Model Results (Scale):
# 
#                           estimate      se      tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -2.4057  0.1423  -16.9060  316  <.0001  -2.6856  -2.1257  *** 
# Org.fertilizer.typeplant    0.8071  0.1808    4.4634  316  <.0001   0.4513   1.1629  *** 

```

#### lnCVR

```{R}
#| eval: false

mod_lnCVR_ls <- rma(yi = lnCVR,
                   vi = var.lnCVR,
                   mods = ~ Org.fertilizer.type,
                   scale = ~ Org.fertilizer.type,
                   data = dat,
                   test = "t",
                   method = "REML"
                   )
summary(mod_lnCVR_ls)
# Location-Scale Model (k = 318; tau^2 estimator: REML)
# 
#    logLik   deviance        AIC        BIC       AICc   
# -437.4012   874.8024   882.8024   897.8254   882.9310   
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 795.3961, p-val < .0001
# 
# Test of Location Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 1.0782, p-val = 0.2999
# 
# Model Results (Location):
# 
#                           estimate      se    tval   df    pval    ci.lb   ci.ub      
# intrcpt                     0.2374  0.0616  3.8525  316  0.0001   0.1161  0.3586  *** 
# Org.fertilizer.typeplant    0.1021  0.0984  1.0384  316  0.2999  -0.0914  0.2957      
# 
# Test of Scale Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 10.1669, p-val = 0.0016
# 
# Model Results (Scale):
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -1.5168  0.2481  -6.1136  316  <.0001  -2.0049  -1.0286  *** 
# Org.fertilizer.typeplant    1.0459  0.3280   3.1886  316  0.0016   0.4005   1.6913   ** 

```

#### lnVR

```{R}
#| eval: false

mod_lnVR_ls <- rma(yi = lnVR,
                   vi = var.lnVR,
                   mods = ~ Org.fertilizer.type,
                   scale = ~ Org.fertilizer.type,
                   data = dat,
                   test = "t",
                   method = "REML"
                   )
summary(mod_lnVR_ls)
# Location-Scale Model (k = 318; tau^2 estimator: REML)
# 
#    logLik   deviance        AIC        BIC       AICc   
# -415.2505   830.5010   838.5010   853.5240   838.6296   
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 651.1955, p-val < .0001
# 
# Test of Location Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 4.7895, p-val = 0.0294
# 
# Model Results (Location):
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub    
# intrcpt                     0.0417  0.0457   0.9106  316  0.3632  -0.0483   0.1317    
# Org.fertilizer.typeplant   -0.1906  0.0871  -2.1885  316  0.0294  -0.3620  -0.0193  * 
# 
# Test of Scale Coefficients (coefficient 2):
# F(df1 = 1, df2 = 316) = 11.9043, p-val = 0.0006
# 
# Model Results (Scale):
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -2.8652  0.6270  -4.5699  316  <.0001  -4.0987  -1.6316  *** 
# Org.fertilizer.typeplant    2.3044  0.6679   3.4503  316  0.0006   0.9903   3.6185  *** 

```

The models suggest that plant-based fertilisers are associated with both lower mean effects and substantially greater variability across studies.
In particular, the scale component indicates that effect sizes are markedly more dispersed for plant-based fertilisers across all three variability metrics.
However, these models treat all effect sizes as independent and do not account for the multilevel structure of the data.
<!-- Should we add the estimated values? -->

### B. Heteroscedastic meta-regression

Heteroscedastic meta-regression models allow residual variance to differ across levels of a moderator while incorporating a multilevel random-effects structure.
By relaxing the assumption of constant heterogeneity, these models permit different levels of a moderator to be associated with different amounts of residual variability.
Although this approach does not constitute a full multilevel location–scale model, it provides a useful intermediate step toward more flexible variance modelling.
In particular, it enables the examination of moderator effects on both the mean effect size and the residual heterogeneity while accounting for the multilevel structure of the data.

In the present heteroscedastic meta-regression, the random-effects structure `~ Org.fertilizer.type | effect_size_ID` allows residual heterogeneity at the effect-size level to differ across fertiliser types.
By specifying `struct = "DIAG"`, the model estimates fertiliser-type-specific variances without assuming covariance between categories, thereby relaxing the assumption of homogeneous residual variability while retaining a parsimonious specification.

::: callout-tip
We used this heteroscedastic meta-regression models for figures 3A and 3B in the main text.
:::

#### lnRR

```{r}
#| eval: false

mod_lnRR_mr2 <- rma.mv(lnRR,
                       var.lnRR,
                       mods = ~ Org.fertilizer.type,
                       random = list(~ 1 | study_ID,
                                     ~ Org.fertilizer.type | effect_size_ID),
                       data = dat,
                       test = "t",
                       method = "REML",
                       struct =  'DIAG'
                       )
summary(mod_lnRR_mr2)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -105.4125   210.8250   220.8250   239.6037   221.0185   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed    factor 
# sigma^2    0.0608  0.2466     36     no  study_ID 
# 
# outer factor: effect_size_ID      (nlvls = 318)
# inner factor: Org.fertilizer.type (nlvls = 2)
# 
#             estim    sqrt  k.lvl  fixed   level 
# tau^2.1    0.0666  0.2581    134     no  animal 
# tau^2.2    0.0937  0.3061    184     no   plant 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 13892.9906, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 5.9453, p-val = 0.0153
# 
# Model Results:
# 
#                           estimate      se     tval   df    pval    ci.lb    ci.ub      
# intrcpt                    -0.2679  0.0537  -4.9852  316  <.0001  -0.3736  -0.1622  *** 
# Org.fertilizer.typeplant   -0.1399  0.0574  -2.4383  316  0.0153  -0.2528  -0.0270    * 

```

```{r}
#| echo: false
#| fig-align: center

orchard_mr_lnRR <- read_cached_artifact("orchard_mr_lnRR.rds")
orchard_mr_lnRR

```

#### lnCVR

```{r}
#| eval: false
#| fig-align: center

mod_lnCVR_mr2 <- rma.mv(lnCVR,
                       var.lnCVR,
                       mods = ~ Org.fertilizer.type,
                       random = list(~ 1 | study_ID,
                                     ~ Org.fertilizer.type | effect_size_ID),
                       data = dat,
                       test = "t",
                       method = "REML",
                       struct =  'DIAG'
                       )
summary(mod_lnCVR_mr2)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -432.6704   865.3408   875.3408   894.1195   875.5343   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed    factor 
# sigma^2    0.0555  0.2356     36     no  study_ID 
# 
# outer factor: effect_size_ID      (nlvls = 318)
# inner factor: Org.fertilizer.type (nlvls = 2)
# 
#             estim    sqrt  k.lvl  fixed   level 
# tau^2.1    0.1876  0.4332    134     no  animal 
# tau^2.2    0.5302  0.7282    184     no   plant 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 795.3961, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 0.5057, p-val = 0.4775
# 
# Model Results:
# 
#                           estimate      se    tval   df    pval    ci.lb   ci.ub      
# intrcpt                     0.2989  0.0812  3.6818  316  0.0003   0.1392  0.4586  *** 
# Org.fertilizer.typeplant    0.0845  0.1189  0.7112  316  0.4775  -0.1494  0.3185      

```

```{r}
#| echo: false
#| fig-align: center

orchard_mr_lnCVR <- read_cached_artifact("orchard_mr_lnCVR.rds")
orchard_mr_lnCVR

```

#### lnVR

```{r}
#| eval: false

mod_lnVR_mr2 <- rma.mv(lnVR,
                       var.lnVR,
                       mods = ~ Org.fertilizer.type,
                       random = list(~ 1 | study_ID,
                                     ~ Org.fertilizer.type | effect_size_ID),
                       data = dat,
                       test = "t",
                       method = "REML",
                       struct =  'DIAG'
                       )
summary(mod_lnVR_mr2)
# Multivariate Meta-Analysis Model (k = 318; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -409.8610   819.7221   829.7221   848.5008   829.9156   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed    factor 
# sigma^2    0.0588  0.2425     36     no  study_ID 
# 
# outer factor: effect_size_ID      (nlvls = 318)
# inner factor: Org.fertilizer.type (nlvls = 2)
# 
#             estim    sqrt  k.lvl  fixed   level 
# tau^2.1    0.0291  0.1707    134     no  animal 
# tau^2.2    0.5092  0.7136    184     no   plant 
# 
# Test for Residual Heterogeneity:
# QE(df = 316) = 651.1955, p-val < .0001
# 
# Test of Moderators (coefficient 2):
# F(df1 = 1, df2 = 316) = 1.9172, p-val = 0.1671
# 
# Model Results:
# 
#                           estimate      se     tval   df    pval    ci.lb   ci.ub    
# intrcpt                     0.0666  0.0713   0.9329  316  0.3516  -0.0738  0.2069    
# Org.fertilizer.typeplant   -0.1526  0.1102  -1.3846  316  0.1671  -0.3695  0.0643    

```

```{r}
#| echo: false
#| fig-align: center

orchard_mr_lnVR <- read_cached_artifact("orchard_mr_lnVR.rds")
orchard_mr_lnVR

```

In the heteroscedastic meta-regression for lnRR, residual heterogeneity at the effect-size level differed between fertiliser types, with larger heterogeneity for plant-based fertilisers ($\tau^2 = 0.094$) than for animal-based fertilisers ($\tau^2 = 0.067$).
The location component indicated a significantly lower mean lnRR for plant-based fertilisers relative to animal-based fertilisers (estimate = −0.14, 95% CI = [-0.25, -0.03]).
Thus, even after accounting for multilevel structure, plant-based fertilisers were associated with both lower mean effects and moderately greater residual heterogeneity.

For lnCVR, allowing residual heterogeneity to vary across fertiliser types revealed substantial differences in heterogeneity magnitude.
Effect-size–level heterogeneity was markedly larger for plant-based fertilisers ($\tau^2 = 0.54$) than for animal-based fertilisers ($\tau^2 = 0.19$).
However, the location effect of fertiliser type on mean lnCVR remained small and statistically unsupported (estimate = 0.08, 95% CI = [-0.15, 0.32]), indicating no consistent difference in average relative variability between fertiliser categories.

A similar pattern was observed for lnVR.
Residual heterogeneity was considerably larger for plant-based fertilisers ($\tau^2 = 0.51$) than for animal-based fertilisers ($\tau^2 = 0.03$), suggesting pronounced differences in the dispersion of absolute variability across fertiliser types.
In contrast, the location effect on mean lnVR was not statistically supported (estimate = −0.15, 95% CI = [-0.37, 0.06]), indicating weak evidence for systematic differences in average absolute variability.
::::

## brms

`brms` provides a flexible framework for fitting location–scale (LS) models, in which the mean structure of the response variable (the location component) and the residual variability around that mean (the scale component) are modelled simultaneously.
This allows us to examine not only how predictors affect average effect sizes, but also how they influence heterogeneity.

In this tutorial, we focus primarily on a location–scale meta-regression model with fixed effects in the scale component.
This model represents the most stable and interpretable formulation and serves as the main working example throughout the `brms` section.

For completeness, we also briefly introduce two more complex extensions:

A.  Location–scale meta-regression with fixed effects in the scale component.\
    This is the primary model used in this tutorial and the only one for which we provide full interpretation of results.

B.  Location–scale meta-regression with random effects in the scale component.\
    This model allows heterogeneity to vary across higher-level grouping factors (for example, studies), but may be difficult to fit reliably in practice.

C.  Location–scale meta-regression with correlated random effects in the location and scale components.\
    This formulation allows the mean and heterogeneity to be evolutionarily or structurally linked, but often suffers from convergence issues in applied meta-analytic datasets.

Models B and C are included to illustrate the range of model structures that are theoretically possible in `brms`.
We provide one concrete example of each, but we do not interpret their results in detail, as these models frequently exhibit convergence problems and require a lot of data.
If you are interested in more detail, please check Nakagawa et al. (2025[a](https://doi.org/10.1111/gcb.70204), [b](https://doi.org/10.1111/2041-210x.70203)).

It is worth noting that, in the models presented here, we specify the scale component using the `sigma ~` syntax in `brms`.
This choice reflects the use of a Gaussian response distribution, for which `sigma` represents the residual standard deviation.
For other response distributions, the scale component may be parametrised differently in `brms` (for example, using `shape` or `phi` - see [here](https://ayumi-495.github.io/Eco_location-scale_model/#specifying-the-scale-component)), and the corresponding model formula should be adapted accordingly.

::: callout-note
### Keep in your mind…
In the location–scale models by `brms`, dispersion parameters are modelled on the log standard deviation scale ($\ln(\sigma)$). All coefficients for the scale component should be interpreted as effects on the log standard deviation, not on the variance.

### Practical guidance

Models with random effects in the scale component, and especially models with correlated random effects across location and scale, require:

-   Sufficient numbers of studies

-   Adequate replication within studies

-   Careful prior specification

-   Thorough convergence diagnostics

For these reasons, we treat Models A and B as examples.
We recommend that readers begin with a simpler location–scale model that uses fixed effects in the scale part, and only add complexity if the data clearly support it, and also, you have enough data.
:::

::: panel-tabset
### A. Only fixed effects in scale component

In this model, we include fixed effects in both location and scale components, but no random effects in the scale component (of course, we can include any random effects on location component).

Between-study heterogeneity in mean effect sizes is captured through a random intercept for `study_ID` in the location component, and residual heterogeneity is allowed to vary systematically with predictors via fixed effects in the scale component.
No random variation in residual variance across studies is assumed.

#### lnRR

```{r}
#| eval: false

form_ls_ma1 <- bf(lnRR ~ 1 + Org.fertilizer.type +
                    (1|study_ID) + # this is u_l (the between-study effect)
                    (1|gr(effect_size_ID, cov = vcv)), # this is m (sampling error)
                  sigma ~ 1 +  Org.fertilizer.type # fixed effects only in scale component
                  )

# generate default prior
prior_ls_ma1 <- default_prior(form_ls_ma1, 
                              data=dat, 
                              data2=list(vcv=vcv),
                              family=gaussian())

prior_ls_ma1$prior[5] = "constant(1)" # fixing the variance to 1 

# fitting model
fit_ls_ma1 <- brm(
  formula = form_ls_ma1,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_ls_ma1,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_ls_ma1)
# Family: gaussian 
#   Links: mu = identity; sigma = log 
# Formula: lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#          sigma ~ 1 + Org.fertilizer.type
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.45      0.27     0.02     1.02 1.00      865     2021
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.26      0.04     0.19     0.35 1.00     2342     3505
# 
# Regression Coefficients:
#                                Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept                         -0.26      0.06    -0.38    -0.16 1.00     2437     3607
# sigma_Intercept                   -1.30      0.07    -1.45    -1.15 1.00     3773     3686
# Org.fertilizer.typeplant          -0.15      0.06    -0.27    -0.03 1.00     5617     4995
# sigma_Org.fertilizer.typeplant     0.17      0.09    -0.01     0.35 1.00     9838     4318

```

```{r}
#| echo: false
#| fig-align: center

fit_ls_ma1 <- read_cached_artifact("fit_ls_ma1.rds")

p3 <- visualize_fixed_effects(fit_ls_ma1)
p4 <- visualize_random_effects2(fit_ls_ma1)
```

For lnRR, the location component indicated a negative average effect under animal-based fertilisers ($\beta^{(l)}_{\text{animal}} = -0.26$, 95% credible interval CrI = [-0.38, -0.16]), implying that organic yields are, on average, lower than conventional yields under animal-based fertilisation.
The contrast coefficient for fertiliser type was also negative ($\beta^{(l)}_{\text{plant-animal}} = -0.15$, 95% CrI = [-0.27, -0.03]), indicating an additional yield penalty under plant-based fertilisers relative to animal-based fertilisers.
On the response-ratio scale, these estimates correspond to an expected organic yield of $\exp(\beta_0) = \exp(-0.26) \approx 0.77$, or about 23% lower than conventional yields, under animal-based fertilisers (the reference category).
For plant-based fertilisers, the predicted response ratio is given by $\exp(\!\beta_0 + \beta_{\text{plant-animal}}\!) = \exp(-0.26 - 0.15) \approx 0.66$, corresponding to organic yields that are approximately 34% lower than conventional yields.
The scale intercept for the animal-based reference category was -1.30, which corresponds to a residual standard deviation of $\exp(−1.30) \approx 0.27$ on the lnRR effect-size scale. Importantly, this intercept describes residual heterogeneity among lnRR effect sizes and does not represent a direct organic-conventional difference in yield variability.
The scale component, modelled on the log-$\sigma$ scale, suggested a tendency towards greater residual heterogeneity under plant-based fertilisers ($\beta^{(s)}_{\text{plant-animal}} = 0.17$, 95% CrI = [-0.01, 0.35]).


This corresponds to an approximate 19% increase in residual heterogeneity for plant-based systems ($\exp(0.17) - 1 \approx 0.19$), although the credible interval still overlapped zero, indicating moderate uncertainty in the magnitude of this effect.

#### lnCVR

```{r}
#| eval: false

form_ls_ma2 <- bf(lnCVR ~ 1 + Org.fertilizer.type +
                    (1|study_ID) + # this is u_l (the between-study effect)
                    (1|gr(effect_size_ID, cov = vcv)), # this is m (sampling error)
                  sigma ~ 1 +  Org.fertilizer.type
                  )

# generate default priors
prior_ls_ma2 <- default_prior(form_ls_ma2, 
                              data=dat, 
                              data2=list(vcv=vcv),
                              family=gaussian())

prior_ls_ma2$prior[5] = "constant(1)" # meta-analysis assumes

# fitting model
fit_ls_ma2 <- brm(
  formula = form_ls_ma2,
  data = dat,
  data2 = list(vcv=vcv),
  chains = 2,
  cores = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior_ls_ma2,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(fit_ls_ma2)
# Family: gaussian 
#   Links: mu = identity; sigma = log 
# Formula: lnCVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + (1 | gr(effect_size_ID, cov = vcv)) 
#          sigma ~ 1 + Org.fertilizer.type
#    Data: dat (Number of observations: 318) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect_size_ID (Number of levels: 318) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_ID (Number of levels: 36) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.26      0.08     0.12     0.42 1.00     1557     2388
# 
# Regression Coefficients:
#                                Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept                          0.30      0.09     0.13     0.47 1.00     5124     4713
# sigma_Intercept                   -0.84      0.13    -1.10    -0.60 1.00     1309     2382
# Org.fertilizer.typeplant           0.08      0.12    -0.16     0.31 1.00     4922     5061
# sigma_Org.fertilizer.typeplant     0.51      0.18     0.17     0.86 1.00     1716     3396
# 
# Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
# and Tail_ESS are effective sample size measures, and Rhat is the potential
# scale reduction factor on split chains (at convergence, Rhat = 1).

```

```{r}
#| echo: false
#| fig-align: center

fit_ls_ma2 <- read_cached_artifact("fit_ls_ma2.rds")

p3 <- visualize_fixed_effects(fit_ls_ma2)
p4 <- visualize_random_effects2(fit_ls_ma2)
p3 + p4

```

For lnCVR, the location component indicated that organic systems tend to exhibit higher mean-corrected variability under animal-based fertilisers ($\beta^{(l)}_{\text{animal}} = 0.30$, 95% CrI = [0.13, 0.47]), corresponding to an average CV ratio of $\exp(0.30) \approx 1.35$, roughly 35% higher variability in organic relative to conventional systems.

The contrast coefficient for fertiliser type in the location component was small and highly uncertain ($\beta^{(l)}_{\text{plant-animal}} = 0.08$, 95% CrI = [-0.16, 0.31]), suggesting that fertiliser category does not systematically shift the average mean-corrected within-group-variability contrast between organic and conventional systems.

In contrast, the scale component showed a larger residual-scale component for plant-based fertilisers.
Residual heterogeneity was substantially larger under plant-based fertilisers ($\beta^{(s)}_{\text{plant-animal}} = 0.51$, 95% CrI = [0.17, 0.86]), corresponding to a multiplicative increase in residual variability of approximately 67% ($\exp(0.51) \approx 1.67$).
Under this model, the implied residual standard deviation for lnCVR under plant-based fertilisers is $\sigma \approx \exp(-0.84 + 0.51) \approx 0.72$.

Thus, while the expected magnitude of the mean-corrected within-group-variability contrast (lnCVR) is broadly similar across fertiliser categories, the represented comparisons have a wider residual spread under plant-based fertilisers.
This residual-scale contrast concerns consistency within the analysed studies; it does not by itself establish poorer transferability to a new study.

<!-- #### lnVR -->

<!-- not converged!! -->

<!-- ```{r} -->

<!-- #| eval: false -->

<!-- form_ls_ma3 <- bf(lnVR ~ 1 + Org.fertilizer.type + -->

<!--                     (1|study_ID) + # this is u_l (the between-study effect) -->

<!--                     (1|gr(effect_size_ID, cov = vcv)), # this is m (sampling error) -->

<!--                   sigma ~ 1 +  Org.fertilizer.type -->

<!--                   ) -->

<!-- # generate default priors -->

<!-- prior_ls_ma3 <- default_prior(form_ls_ma3, -->

<!--                               data=dat, -->

<!--                               data2=list(vcv=vcv), -->

<!--                               family=gaussian()) -->

<!-- prior_ls_ma3$prior[5] = "constant(1)" # meta-analysis assumes -->

<!-- # fitting model -->

<!-- fit_ls_ma3 <- brm( -->

<!--   formula = form_ls_ma3, -->

<!--   data = dat, -->

<!--   data2 = list(vcv=vcv), -->

<!--   chains = 2, -->

<!--   cores = 2, -->

<!--   iter = 6000, -->

<!--   warmup = 3000, -->

<!--   prior = prior_ls_ma3, -->

<!--   control = list(adapt_delta=0.95, max_treedepth=15) -->

<!--   ) -->

<!-- summary(fit_ls_ma3) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- #| eval: false -->

<!-- #| echo: false -->

<!-- #| fig-align: center -->

<!-- fit_ls_ma3 <- readRDS(here::here("Rdata", "fit_ls_ma3.rds")) -->

<!-- p3 <- visualize_fixed_effects(fit_ls_ma3) -->

<!-- p4 <- visualize_random_effects2(fit_ls_ma3) -->

<!-- p3 + p4 -->

<!-- ``` -->

### B and C. Random effects/Correlated random effects in location and scale components

The models presented here extend the basic location–scale meta-regression by allowing study-level heterogeneity to operate not only on the mean effect size, but also on the residual variance.

These models are included for completeness and to illustrate the flexibility of `brms`.
In practice, they are often difficult to fit reliably and should be treated as advanced extensions rather than default choices.
Accordingly, we provide example code but do not interpret the results in detail.

For transparency, examples in which the models were run but failed to converge are also documented in this [online tutorial](https://itchyshin.github.io/location-scale_meta-analysis/).
You can also find more examples of complex location-scale models in another [online tutorial](https://ayumi-495.github.io/Eco_location-scale_model/), though these are not specific to meta-analysis.

**1. with random effects in scale component**

In the model with the random effects in the scale component, we can include random effects in the scale component, allowing residual heterogeneity to vary across studies.
If we want to specify such a model, we can do so as follows:

```{r}
#| eval: false

sigma ~ 1 + Org.fertilizer.type + (1 | study_ID)

```

This specification allows the residual variance to vary across studies, in addition to allowing study-level variation in the mean effect size via the location component.
Conceptually, the location component models how the mean effect size varies with the moderator and across studies.
The scale component models how the amount of unexplained variability differs across studies and fertiliser types.

Random effects in the location and scale components are assumed to be independent in this formulation.

**2. Correlated random effects in location and scale components**

The next model further extends the previous formulation by allowing correlated random effects between the location and scale components.
For example, if we want to consider potential correlations between study-level deviations in mean effect sizes and deviations in residual variance, we can specify the model as follows:

```{r}
#| eval: false

(1 | p | study_ID)

```

This specification allows the model to estimate the correlation between study-level deviations in the mean effect size and deviations in residual variance.

For example, a negative correlation would indicate that studies with larger mean effects tend to show lower residual variability, whereas a positive correlation would suggest the opposite.
Such relationships are conceptually appealing and biologically interpretable in some contexts.

However, this model is computationally demanding and highly sensitive to prior specification and data structure.
In applied meta-analyses, it frequently exhibits convergence problems or weakly identified correlation parameters.
:::

## glmmTMB

In contrast to `metafor`, `glmmTMB` allows us to include random effects in both location and scale components - we demonstrate how to fit a location-scale meta-regression model using `glmmTMB`.

But it should be mentioned that `glmmTMB` does not support correlated random effects between location and scale components, so the third model presented above cannot be implemented using `glmmTMB`.

Again, the same as `brms`, we focus on a location–scale meta-regression model with fixed effects in the scale component.

::: callout-note
### Keep in your mind…
In the location–scale models by `glmmTMB`, dispersion parameters are modelled on the log standard deviation scale ($\ln(\sigma)$). All coefficients for the scale component should be interpreted as effects on the log standard deviation, not on the variance.
:::

### A. Only fixed effects in scale component

In `glmmTMB`, the location–scale structure is implemented by separating the model into a *conditional* component and a *dispersion* component.
The conditional model specifies the mean structure of the response (location), while the dispersion model specifies how the residual variance changes as a function of predictors (scale).
Note that the dispersion model in `glmmTMB` does not describe an additional random effect, but rather models heteroscedasticity in the residual variance on the log scale.

As shown below, the location–scale models with `glmmTMB` produce results that match those from `brms` completely.

#### lnRR

```{r}
#| eval: false

dat$effect_size_ID <- factor(dat$effect_size_ID)
dat$g <- factor(1)
vcv <- diag(dat$var.lnRR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnRR1 <- glmmTMB(lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + 
                      equalto(0 + effect_size_ID|g, vcv),
                  dispformula = ~ 1 + Org.fertilizer.type, # scale component
                  data = dat,
                  REML = TRUE)
summary(glmm_lnRR1)
#  Family: gaussian  ( identity )
# Formula:          lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +      effect_size_ID | g, vcv)
# Dispersion:            ~1 + Org.fertilizer.type
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     230.9     249.7    -110.5     220.9       313 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr                                            
#  study_ID (Intercept)      0.0608129 0.24660                                                  
#  g        effect_size_ID33 0.0047213 0.06871                                                  
#           effect_size_ID34 0.0426425 0.20650  0.00                                            
#           effect_size_ID35 0.0698313 0.26426  0.00 0.00                                       
#           effect_size_ID36 0.0798567 0.28259  0.00 0.00 0.00                                  
#           effect_size_ID37 0.0837435 0.28938  0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.0095870 0.09791  0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.0091520 0.09567  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.0363323 0.19061  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.0331452 0.18206  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.0012338 0.03513  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                         NA      NA                                                  
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -0.26791    0.05379  -4.981 6.33e-07 ***
# Org.fertilizer.typeplant -0.13991    0.05897  -2.373   0.0177 *  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Dispersion model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -1.35445    0.07922 -17.097   <2e-16 ***
# Org.fertilizer.typeplant  0.17055    0.09928   1.718   0.0858 . 

```

#### lnCVR

```{r}
#| eval: false
library(glmmTMB)
dat$effect_size_ID <- factor(dat$effect_size_ID)
dat$g <- factor(1)
vcv <- diag(dat$var.lnCVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnCVR1 <- glmmTMB(lnCVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      dispformula = ~ 1 + Org.fertilizer.type,
                      data = dat,
                      REML = TRUE)
summary(glmm_lnCVR1)
#  Family: gaussian  ( identity )
# Formula:          lnCVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +      effect_size_ID | g, vcv)
# Dispersion:             ~1 + Org.fertilizer.type
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     885.5     904.3    -437.7     875.5       313 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr                                            
#  study_ID (Intercept)      0.05551  0.2356                                                   
#  g        effect_size_ID33 0.11583  0.3403                                                   
#           effect_size_ID34 0.29264  0.5410   0.00                                            
#           effect_size_ID35 0.31983  0.5655   0.00 0.00                                       
#           effect_size_ID36 0.32986  0.5743   0.00 0.00 0.00                                  
#           effect_size_ID37 0.33374  0.5777   0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.12070  0.3474   0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.12026  0.3468   0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.28633  0.5351   0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.28315  0.5321   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.25123  0.5012   0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                         NA      NA                                                  
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)               0.29891    0.08215   3.638 0.000274 ***
# Org.fertilizer.typeplant  0.08455    0.11985   0.705 0.480515    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Dispersion model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -0.8366     0.1303   -6.422 1.35e-10 ***
# Org.fertilizer.typeplant  0.5194     0.1777    2.923  0.00346 ** 

```

<!-- #### lnVR -->

```{r}
#| eval: false
#| echo: false

dat$effect_size_ID <- factor(dat$effect_size_ID)
dat$g <- factor(1)
vcv <- diag(dat$var.lnVR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnVR1 <- glmmTMB(lnVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      dispformula = ~ 1 + Org.fertilizer.type,
                      data = dat,
                      REML = TRUE)
summary(glmm_lnVR1)
#  Family: gaussian  ( identity )
# Formula:          lnVR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +      effect_size_ID | g, vcv)
# Dispersion:            ~1 + Org.fertilizer.type
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     901.0     919.8    -445.5     891.0       313 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr                                            
#  study_ID (Intercept)      0.0872038 0.29530                                                  
#  g        effect_size_ID33 0.0047213 0.06871                                                  
#           effect_size_ID34 0.0426425 0.20650  0.00                                            
#           effect_size_ID35 0.0698313 0.26426  0.00 0.00                                       
#           effect_size_ID36 0.0798567 0.28259  0.00 0.00 0.00                                  
#           effect_size_ID37 0.0837435 0.28938  0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.0095870 0.09791  0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.0091520 0.09567  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.0363323 0.19061  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.0331452 0.18206  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.0012338 0.03513  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                         NA      NA                                                  
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)
# (Intercept)               0.09693    0.08462   1.146    0.252
# Org.fertilizer.typeplant -0.19232    0.14428  -1.333    0.183
# 
# Dispersion model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -0.51541    0.07028  -7.334 2.24e-13 ***
# Org.fertilizer.typeplant  0.75408    0.08808   8.561  < 2e-16 ***

```

### B. Random effects in location and scale components

If you want to include random effects in the scale component, you just need to insert them to the `dispformula` part of the model.

```{r}
#| eval: false

dat$effect_size_ID <- factor(dat$effect_size_ID)
dat$g <- factor(1)
vcv <- diag(dat$var.lnRR)
row.names(vcv) <- colnames(vcv) <- dat$effect_size_ID

glmm_lnRR2 <- glmmTMB(lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + 
                        equalto(0 + effect_size_ID|g, vcv),
                      dispformula = ~ 1 + Org.fertilizer.type + (1 | study_ID),
                      data = dat,
                      REML = TRUE)
summary(glmm_lnRR2)
#  Family: gaussian  ( identity )
# Formula:          lnRR ~ 1 + Org.fertilizer.type + (1 | study_ID) + equalto(0 +  
#     effect_size_ID | g, vcv)
# Dispersion:            ~1 + Org.fertilizer.type + (1 | study_ID)
# Data: dat
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#     177.4     199.9     -82.7     165.4       312 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name             Variance  Std.Dev. Corr                                            
#  study_ID (Intercept)      0.0570683 0.23889                                                  
#  g        effect_size_ID33 0.0047213 0.06871                                                  
#           effect_size_ID34 0.0426425 0.20650  0.00                                            
#           effect_size_ID35 0.0698313 0.26426  0.00 0.00                                       
#           effect_size_ID36 0.0798567 0.28259  0.00 0.00 0.00                                  
#           effect_size_ID37 0.0837435 0.28938  0.00 0.00 0.00 0.00                             
#           effect_size_ID39 0.0095870 0.09791  0.00 0.00 0.00 0.00 0.00                        
#           effect_size_ID40 0.0091520 0.09567  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_size_ID42 0.0363323 0.19061  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_size_ID44 0.0331452 0.18206  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_size_ID45 0.0012338 0.03513  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...              ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                         NA      NA                                                  
# Number of obs: 318, groups:  study_ID, 36; g, 1
# 
# Dispersion model:
#  Groups   Name        Variance Std.Dev.
#  study_ID (Intercept) 0.06081  0.2466  
#  Residual                  NA      NA  
# Number of obs: 318, groups:  study_ID, 36
# 
# Conditional model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)              -0.26476    0.04887  -5.417 6.05e-08 ***
# Org.fertilizer.typeplant -0.11847    0.04545  -2.607  0.00914 ** 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Dispersion model:
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)               -1.6225     0.1147  -14.14  < 2e-16 ***
# Org.fertilizer.typeplant   0.4022     0.1406    2.86  0.00424 ** 

```

# Extension: Where does heterogeneity occur?

This extension addresses a different question from the average effect: whether category-specific heterogeneity is attributable mainly to differences **among studies** or to differences **among comparisons within studies**. We distinguish these components because a wider residual distribution supports a statement about consistency among represented comparisons, whereas a wider study distribution is the component relevant to variation in a new study's mean effect.

The extension uses the cereal subset with plant- and animal-based organic fertilisers. The full analysis contains 318 effects from 36 studies. Six studies contribute both fertiliser categories (86 effects); a matched subset excluding those studies contains 232 effects from 30 studies. The latter restriction makes fertiliser category constant within every study, which is required for the direct study-scale formulation below.

```{r}
#| label: study-level-artifact-setup
#| include: false

study_level_artifacts <- read_cached_artifact("study_level_artifacts.rds")
stopifnot(
  identical(
    unname(unlist(study_level_artifacts$counts)),
    c(318L, 36L, 6L, 86L, 232L, 30L)
  )
)
study_level_ratios <- study_level_artifacts$ratios
study_level_predictions <- study_level_artifacts$predictions
study_level_diagnostics <- study_level_artifacts$diagnostics
study_level_frequentist <- study_level_artifacts$frequentist
matched_data <- study_level_artifacts$matched_data
blsmeta_historical <- utils::read.csv(
  "data/blsmeta_sensitivity_verified_outputs.csv",
  stringsAsFactors = FALSE
)
stopifnot(
  nrow(blsmeta_historical) == 4L,
  identical(sort(blsmeta_historical$response), c("lnCVR", "lnCVR", "lnRR", "lnRR")),
  all(blsmeta_historical$convergence_status == "failed_convergence"),
  all(blsmeta_historical$source_commit == "1b759ac")
)

format_ratio <- function(response_name, component_name) {
  row <- study_level_ratios[
    study_level_ratios$response == response_name &
      study_level_ratios$component == component_name &
      study_level_ratios$model == "extended_primary",
    ]
  stopifnot(nrow(row) == 1L)
  sprintf("%.2f [%.2f, %.2f]", row$estimate, row$lower, row$upper)
}
```

## Starting comparator: residual location--scale model

The residual location--scale model is the useful starting comparator.  It
allows fertiliser category to change the residual SD among effect sizes within
studies, while retaining one common SD for the distribution of study means:

$$
y_{ij}=\mu_{g[ij]}+u_j+e_{ij}+m_{ij},\qquad
u_j\sim\mathcal N(0,\tau^2),\quad
e_{ij}\sim\mathcal N(0,\sigma_{g[ij]}^2),\quad
m_{ij}\sim\mathcal N(0,v_{ij}).
$$

Here $\tau$ is between-study heterogeneity and $\sigma_g$ is residual
within-study heterogeneity.  A category difference in $\sigma_g$ is therefore
not, by itself, a difference in transferability to a new study.  The original
residual model is available in `metafor`, `glmmTMB`, and `brms` through a
category-specific random-effect or dispersion/residual-scale component.

## General multilevel location--scale model

For the extension, the two variance components receive separate scale
sub-models:

$$
\begin{aligned}
y_{ij} &= \mathbf{x}_{ij}^{\mathsf T}\boldsymbol\beta+u_j+e_{ij}+m_{ij},\\
u_j &\sim \mathcal N(0,\tau_j^2),\qquad
e_{ij}\sim\mathcal N(0,\sigma_{ij}^2),\qquad
\mathbf m\sim\mathcal N(\mathbf0,\mathbf V),\\
\log(\tau_j) &= \mathbf z_j^{\mathsf T}\boldsymbol\gamma_u,\qquad
\log(\sigma_{ij})=\mathbf w_{ij}^{\mathsf T}\boldsymbol\gamma_e.
\end{aligned}
$$

The location, study-scale, and residual-scale moderators need not be the same.
A study-scale predictor must be constant within study; a residual-scale
predictor may vary among effect sizes.  With correlated sampling errors, the
scalar $v_{ij}$ is replaced by the sampling covariance matrix $\mathbf V$.

## Categorical random coefficients are a location--scale special case

For animal ($A$) and plant ($P$) fertiliser, category-specific study random
coefficients can be written as

$$
y_{ij}=\beta_A I(g_{ij}=A)+\beta_P I(g_{ij}=P)
+u_{jA}I(g_{ij}=A)+u_{jP}I(g_{ij}=P)+e_{ij}+m_{ij},
$$

with $\operatorname{SD}(u_{jA})=\tau_A$ and
$\operatorname{SD}(u_{jP})=\tau_P$.  This is both a categorical random-slope
model and a heteroscedastic random-effects model.  Its diagonal version is
equivalent to the categorical scale regression

$$
\log(\tau_g)=\gamma_0+\gamma_1I(g=P),
$$

so that $\tau_A=\exp(\gamma_0)$ and
$\tau_P=\exp(\gamma_0+\gamma_1)$.  Direct scale regression is more general:
it readily accommodates continuous predictors or several predictors in a
scale sub-model.

Under conventional intercept--slope coding, in contrast, the animal study
effect is $u_{0j}$ and the plant study effect is $u_{0j}+u_{1j}$.  Thus

$$
\operatorname{Var}(u_{jP})=\operatorname{Var}(u_{0j})+
\operatorname{Var}(u_{1j})+2\operatorname{Cov}(u_{0j},u_{1j}),
$$

so the random-slope variance alone is **not** the plant-category variance.
Using `0 + fertilizer` makes the two category-specific SDs directly visible.

## Primary full-data categorical model

For each response $y_i$ (lnRR or lnCVR), the primary model uses category-specific study and residual SDs:

$$y_i \sim \mathcal{N}(\mu_{c[i]},\; v_i + \tau_{c[i]}^2 + \sigma_{c[i]}^2),$$

where $v_i$ is the known sampling variance, $\tau_c$ is the study-level SD, and $\sigma_c$ is the residual SD for fertiliser category $c$. The covariance between animal and plant study effects is fixed at zero (a diagonal study covariance). This is a stated modelling restriction, not evidence that the correlation is zero.

This is a categorical random-coefficient model: in `brms` the study term is written `(0 + fertilizer || study_ID)`.  The covariance between category-specific study effects is fixed to zero.  Estimating their correlation would require the six mixed-category studies to carry substantially more information than they do here; the unstructured fit is boundary-prone and is not interpreted.

The reproducible script [`R/study_level_heterogeneity.R`](R/study_level_heterogeneity.R) recalculates lnCVR with `correct = TRUE`, fits the full-data Bayesian model with `brms`, builds posterior ratio and prediction tables, and saves the artifact read above. It deliberately does not download or silently create inputs during rendering.

### Full-data categorical analyses in three packages

The principal calls remain visible here; the accompanying script supplies the
complete data preparation, fit checks, and artifact-writing route.  All use
the full 318 effects from 36 studies.  `metafor` and `glmmTMB` provide
frequentist point-estimate checks, while `brms` provides posterior intervals
and predictions.

```{r}
#| label: study-level-metafor-code
#| eval: false
#| code-fold: true

metafor::rma.mv(
  yi = data[[response]], V = data[[variance]], mods = ~ 0 + fertilizer,
  random = list(~ fertilizer | study_ID, ~ fertilizer | effect_size_ID),
  struct = c("DIAG", "DIAG"), test = "t", method = "REML", data = data
)

# Exploratory only: use "UN" for the first structure to estimate the
# animal--plant study-effect correlation.  With only six mixed studies this
# correlation is boundary-prone and is not interpreted.
```

```{r}
#| label: study-level-glmmtmb-code
#| eval: false
#| code-fold: true

model_data <- transform(
  data,
  study_ID = factor(study_ID),
  effect_size_ID = factor(effect_size_ID),
  g = factor("known_sampling")
)
V <- diag(model_data[[variance]])
dimnames(V) <- list(levels(model_data$effect_size_ID), levels(model_data$effect_size_ID))

glmmTMB::glmmTMB(
  response ~ 0 + fertilizer +
    diag(0 + fertilizer | study_ID) +
    equalto(0 + effect_size_ID | g, V),
  dispformula = ~ 0 + fertilizer,
  REML = TRUE, data = model_data
)
```

```{r}
#| label: study-level-brms-code
#| eval: false
#| code-fold: true

brms::brm(
  brms::bf(
    y | se(sqrt(vi), sigma = TRUE) ~
      0 + fertilizer + (0 + fertilizer || study_ID),
    sigma ~ 0 + fertilizer
  ),
  family = gaussian(), prior = historical_primary_spec$priors,
  chains = 4, iter = 2000, warmup = 1000, seed = 20260937,
  backend = "cmdstanr", control = list(adapt_delta = 0.99), data = data
)
```

The `brms` residual-scale coefficients (`b_sigma_*`) are on the log-SD scale.
They must be exponentiated before forming SD draws or prediction intervals.

```{r}
#| label: study-level-ratios
#| tbl-cap: "Posterior plant-to-animal SD ratios from the full-data primary and prespecified prior-sensitivity models."

knitr::kable(
  transform(
    study_level_ratios[, c("response", "model", "component", "estimate", "lower", "upper")],
    model = ifelse(model == "extended_primary", "primary Exp(2)", "sensitivity Exp(1)"),
    estimate = round(estimate, 2), lower = round(lower, 2), upper = round(upper, 2)
  ),
  row.names = FALSE
)
```

For lnRR, the posterior study-SD ratio is `r format_ratio("lnRR", "study_sd_ratio")`, while the residual-SD ratio is `r format_ratio("lnRR", "residual_sd_ratio")`. For lnCVR, the corresponding ratios are `r format_ratio("lnCVR", "study_sd_ratio")` and `r format_ratio("lnCVR", "residual_sd_ratio")`. Thus, a comparison-specific residual-scale difference should not be relabelled as a study-level transferability difference.

For a log-SD coefficient $\gamma_1$, the plant-to-animal ratio is
$\exp(\gamma_1)$.  Some software instead models log variance,
$\log(\sigma^2)$; for a coefficient on that scale the corresponding SD ratio
is $\exp(\gamma_1/2)$.  This scale distinction is essential when comparing
parameterisations across packages.

```{r}
#| label: study-level-prior-sensitivity
#| eval: false
#| code-fold: true

# Prespecified sensitivity: change only the study-SD prior from Exp(2) to
# Exp(1), refit, and compare posterior ratios plus R-hat/ESS/divergences.
study_sd_prior_sensitivity_spec <- historical_primary_spec
study_sd_prior_sensitivity_spec$priors <- c(
  brms::set_prior("normal(0, 1)", class = "b"),
  brms::set_prior("normal(-1, 1)", class = "b", dpar = "sigma"),
  brms::set_prior("exponential(1)", class = "sd", group = "study_ID")
)
```

The sensitivity is deliberately prespecified and changes no likelihood,
sampling-error treatment, data, or parameterisation.  It assesses prior
influence on the weakly identified study-scale contrast rather than selecting
a prior to reproduce a desired interval.

## Prediction targets

For a latent new comparison, the predictive variance includes the two
heterogeneity components,

$$\tau_c^2+\sigma_c^2.$$

For an observed new comparison, it also includes sampling variation,

$$\tau_c^2 + \sigma_c^2 + v_i.$$

For the mean of $k$ new comparisons from one new study, residual and sampling terms are averaged but the study term is not:

$$
\tau_c^2+\frac{\sigma_c^2}{k}+
\frac{\mathbf1^\mathsf{T}\mathbf V_{\mathrm{new}}\mathbf1}{k^2}.
$$

For independent equal-variance comparisons, the final term reduces to $v/k$.
The script evaluates these scenarios at a **separate median sampling variance
for each response-by-fertiliser category**, rather than one median pooled over
fertiliser categories.  They are explicit prediction scenarios, not generic
claims about every future data set.

```{r}
#| label: study-level-predictions
#| tbl-cap: "Posterior prediction scenarios at category-specific median sampling variance."

knitr::kable(
  transform(
    study_level_predictions,
    median = round(median, 2), q2.5 = round(q2.5, 2), q97.5 = round(q97.5, 2)
  ),
  row.names = FALSE
)
```

## Matched-subset direct study-scale bridge

The following comparison separates changes due to data restriction from changes due to model parameterisation.

| Analysis | Data | Purpose | Interpretation boundary |
|---|---:|---|---|
| Full categorical model | 318 effects; 36 studies | Primary category-specific study and residual SDs | Includes six studies with both categories; diagonal study covariance is assumed. |
| Direct `drmTMB` study-scale model | 232 effects; 30 studies | Places category predictors directly on residual and study SDs | Valid only when category is constant within a study; fitted with ML. |
| `metafor` bridge | Same 232 effects; 30 studies | Tests the impact of the subset using a category-specific diagonal model | Uses REML, so it is a scope bridge rather than an exact estimator match. |

```{r}
#| label: study-level-bridge
#| tbl-cap: "Frequentist SD-ratio comparison for the full data and matched-subset bridge."

knitr::kable(
  transform(
    study_level_frequentist,
    estimate = round(estimate, 2)
  ),
  row.names = FALSE
)
```

When applied to the same matched data scope and when fertiliser category is constant within study, the direct and bridge models provide alternative parameterisations of the same category-specific variance structure. Differences between their numerical estimates can nevertheless arise from ML versus REML estimation and from weakly identified variance components.

```{r}
#| label: study-level-drmtmb-code
#| eval: false
#| code-fold: true

# The script fits this exact model for lnRR, then repeats it for lnCVR.
matched_lnRR <- transform(matched_data, response = lnRR, V = var.lnRR)
drmTMB::drmTMB(
  drmTMB::bf(
    response ~ fertilizer + (1 | study_ID) + drmTMB::meta_V(V = V),
    sigma ~ 0 + fertilizer,
    sd(study_ID) ~ 0 + fertilizer
  ),
  family = gaussian(), data = matched_lnRR, REML = FALSE
)

matched_lnCVR <- transform(matched_data, response = lnCVR, V = var.lnCVR)
# Replace `matched_lnRR` above with `matched_lnCVR` to fit the lnCVR version.
```

`blsmeta` is a direct Bayesian study-scale implementation.  It is intentionally
kept as an optional sensitivity route: it must use the matched subset, with
observations sorted by study and a freshly created numeric study index.  The
fit must not be interpreted unless its convergence checks pass.

```{r}
#| label: study-level-blsmeta-code
#| eval: false
#| code-fold: true

matched_blsmeta <- matched_data[order(matched_data$study_ID,
                                      matched_data$effect_size_ID), ]
matched_blsmeta$es_blsmeta <- seq_len(nrow(matched_blsmeta))
matched_blsmeta$study_blsmeta <- blsmeta::make_study_id(matched_blsmeta$study_ID)

blsmeta::blsmeta(
  yi = lnRR, vi = var.lnRR,
  es_id = es_blsmeta, study_id = study_blsmeta,
  mods = ~ fertilizer,
  mods_scale2 = ~ 0 + fertilizer,  # residual log SD
  mods_scale3 = ~ 0 + fertilizer,  # study log SD
  iter = 20000, warmup = 5000, chains = 4,
  data = matched_blsmeta
)

# Repeat with yi = lnCVR and vi = var.lnCVR, then check split-R-hat and ESS
# after warmup before reporting exp(eta[2] - eta[1]) as the study-SD ratio.
```

`blsmeta` convergence problems are reported as a diagnostic finding, not
silently discarded: the previously attempted matched-subset chains failed the
predeclared convergence criteria.  Its estimates are therefore sensitivity
results only, and neither support nor contradict the primary full-data `brms`
inference.  The primary rendered results do not depend on `blsmeta` being
installed.

### Historical `blsmeta` sensitivity output

The following table documents the previously fitted direct Bayesian study-scale
sensitivity analysis, rather than a new fit made while rendering this tutorial.
Its four ratios are transcribed from historical commit `1b759ac`,
`results/study_level_heterogeneity/model_summary.csv`; the convergence
diagnostics are transcribed from the corresponding `diagnostics.csv`.  The
versioned provenance record is
`data/blsmeta_sensitivity_verified_outputs.csv`.  This explicit record is
necessary because a fresh fit currently requires a compatible `rjags`/JAGS 4.x
toolchain, whereas current Homebrew JAGS 5 is not compatible with `rjags`.

Both `blsmeta` fits **failed** the predeclared convergence criteria.  These values
are shown only to document the requested sensitivity analysis: they are not
claim-bearing results and are not evidence for or against the primary full-data
`brms` conclusions.

```{r}
#| label: study-level-blsmeta-historical-results
#| tbl-cap: "Historical blsmeta sensitivity SD ratios and diagnostics. All rows failed convergence and are documentation only."

blsmeta_display <- data.frame(
  Response = blsmeta_historical$response,
  Component = ifelse(
    blsmeta_historical$component == "residual_sd",
    "Residual SD", "Study SD"
  ),
  `Plant-to-animal SD ratio [95% CrI]` = sprintf(
    "%.2f [%.2f, %.2f]",
    blsmeta_historical$estimate,
    blsmeta_historical$lower,
    blsmeta_historical$upper
  ),
  `Max parameter R-hat` = sprintf("%.3f", blsmeta_historical$max_parameter_rhat),
  `Min parameter ESS` = sprintf("%.0f", blsmeta_historical$min_parameter_ess),
  `Convergence status` = "FAILED: study-ratio target criteria not met",
  check.names = FALSE
)
knitr::kable(blsmeta_display, row.names = FALSE)
```

## Diagnostics and interpretation

```{r}
#| label: study-level-diagnostics
#| tbl-cap: "Diagnostics for the full-data Bayesian primary models."

knitr::kable(
  transform(
    study_level_diagnostics,
    max_rhat = round(max_rhat, 3), min_bulk_ess = round(min_bulk_ess)
  ),
  row.names = FALSE
)
```

The full-data lnRR model addresses study-level variation directly. In contrast, the stronger lnCVR residual-scale contrast describes wider variability among comparisons within the represented studies. It supports a conditional statement about within-study consistency, not an unconditional claim about transferability to new studies. The analysis does not establish a causal mechanism for either component.

The primary Bayesian fits must have $\widehat R\leq1.01$, adequate effective
sample sizes, and no unresolved divergences.  The prior-sensitivity fits are
checked by the same criteria.  `metafor` and `glmmTMB` provide frequentist
cross-checks for the categorical likelihood; they need not agree exactly with
the Bayesian posterior summaries or with each other on the matched subset
because they differ in estimator (REML versus ML) and parameterisation.

Only six studies contain both fertiliser categories.  Consequently, the
animal--plant study-effect correlation is weakly identified, unstructured
covariance models are boundary-prone, and the lnCVR study-scale contrast has
wide uncertainty.  These are limitations of the information in the data, not
evidence for a zero SD or a category-specific causal mechanism.

# Software and package versions

```{r}
#| echo: false

cat(paste(trimws(capture.output(sessionInfo()), which = "right"), collapse = "\n"))

```

# References

-   Lauren C. Ponisio, Leithen K. M'Gonigle, Kevi C. Mace, Jenny Palomino, Perry de Valpine, Claire Kremen; Diversification practices reduce organic to conventional yield gap. Proc Biol Sci. 2015; 282 (1799): 20141396. <https://doi.org/10.1098/rspb.2014.1396>
